---
output: github_document
---

<!-- README.md is generated from README.rmd -- edit README.rmd, then re-knit with
     rmarkdown::render('README.rmd'). The example chunks are EVALUATED against the
     installed ctsem, so install the package before knitting. Write plain markdown
     here: LaTeX (\url{}) and roxygen (#') markup do not survive the conversion. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# ctsem

<!-- badges: start -->
[![R-CMD-check](https://github.com/cdriveraus/ctsem/workflows/R-CMD-check/badge.svg)](https://github.com/cdriveraus/ctsem/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/ctsem)](https://cran.r-project.org/package=ctsem)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/ctsem)](https://cran.r-project.org/package=ctsem)
<!-- badges: end -->

**Hierarchical continuous (and discrete) time dynamic modelling in R.**

ctsem fits state space models -- a stochastic differential or difference equation
for the process, plus a measurement model for the observations -- to longitudinal
data. Panel data with a handful of waves, intensive longitudinal data with hundreds
of observations, or a single long time series all use the same interface. Unequal
and individually varying measurement intervals are handled directly, rather than
being rounded into waves.

You can specify multiple interacting processes, multiple indicators per process
(dynamic factor analysis), higher order dynamics such as damped oscillation, time
varying (within subject) and time invariant (between subject) covariates,
interventions and other exogenous inputs, and state dependent parameters
(nonlinearity). Any parameter may vary over individuals, with its population mean,
variance, and covariate effects estimated jointly. Classic longitudinal models such
as latent growth curves, latent change scores and the cross-lagged panel model are
special cases.

Estimation defaults to maximum likelihood, or maximum a posteriori when priors are
used; Hamiltonian Monte Carlo sampling is available.

If your data are in long format with a subject id column and a time column, you are
ready to start.

## Install

From CRAN:

```r
install.packages('ctsem')
```

Or the development version:

```r
remotes::install_github('cdriveraus/ctsem',
  INSTALL_opts = "--no-multiarch", dependencies = c("Depends", "Imports"))
```

Most models run without compiling anything. Some of the more complex ones compile a
Stan model, which needs a working C++ toolchain.

<details><summary><b>Compile problems on Windows (rstan / Rtools)</b></summary>

Ensure a recent R and a matching Rtools are installed, and check that `rstan`
itself works before suspecting ctsem.

Put this single line in `~/.R/Makevars.win`, deleting any other lines:

```
CXX17FLAGS += -mtune=native -Wno-ignored-attributes -Wno-deprecated-declarations
```

If you see errors like `g++ not found`, install devtools, which brings the
toolchain checks with it:

```r
install.packages('devtools')
```

For anything else compiler related, the Stan forums are the best place to look:
<https://discourse.mc-stan.org/>.

</details>

## A first model

Two coupled processes, densely measured. `milkman_rdm1` in the
[doBy](https://cran.r-project.org/package=doBy) package records daily milk yield and
fat percentage through a full lactation for 53 dairy cows, a median of 230
observations each. That density is what makes individual differences in the
dynamics estimable at all, rather than only differences in level.

```{r firstmodel, results = 'hide'}
library(ctsem)
data(milkman_rdm1, package = 'doBy')

milk <- data.frame(
  cow   = milkman_rdm1$cowno,
  day   = milkman_rdm1$dfc,                     # days since calving
  yield = as.numeric(scale(milkman_rdm1$my)),
  fat   = as.numeric(scale(milkman_rdm1$fatpct)))

model <- ctModel(
  type = 'ct',                        # 'dt' for the discrete time equivalent
  LAMBDA = diag(2),                   # two processes, one indicator each
  manifestNames = c('yield', 'fat'),
  latentNames = c('yield', 'fat'),
  time = 'day',
  id = 'cow')

fit <- ctFit(milk, model)
```

Nothing there is constrained. `LAMBDA` is the factor loading matrix linking latent
processes to observed variables, and its dimensions are what tell ctsem the system
has two processes with one indicator each. Every other matrix -- the auto and cross
effects, the system noise and its correlation, the measurement error, the initial
states -- takes a default and is freely estimated, with individual differences where
ctsem's defaults put them. Scaling the two variables, as above, makes for easier
estimation and more sensible default priors.

```{r firstsummary}
summary(fit)$popmeans
```

Drift names read as `drift_<affected>_<predictor>`, so `drift_yield` is yield's own
auto effect and `drift_yield_fat` is the effect of fat on yield. A negative auto
effect pulls a process back toward its equilibrium and its size sets how fast.
`diff_*` are the system noise standard deviations and their correlation, `mvar*` the
measurement error standard deviations, and `T0m_*` the mean initial states.

The two cross effects are not symmetric: a high fat percentage predicts a
subsequent fall in yield, and the interval for `drift_yield_fat` excludes zero,
while the interval for the reverse effect does not.

```{r firstpopsd}
summary(fit)$popsd
```

Those are the population standard deviations of the parameters that vary by cow.
ctsem's defaults put random effects on the initial state and the observation
intercept; adding `indvarying` to a parameter's specification puts them elsewhere,
on an auto effect or a cross effect, which is what the density of this data buys.

To read the model as equations rather than as code, `ctModelLatex(model)` renders it
to pdf or png.

```{r firstplot, fig.height = 4.4, fig.width = 8, fig.alt = "Observed yield and fat percentage for two cows over one lactation, with the model's expected trajectory and uncertainty band tracking each series."}
ctPredict(fit, subjects = unique(milk$cow)[1:2], plot = TRUE)
```

Points are the observations, lines the model's expectation for each cow, and the
bands its uncertainty. The rise and slow decline in yield is the lactation curve,
which the model is picking up as a long deviation from equilibrium rather than as a
trend term.

```{r firstdiscrete, fig.height = 4.4, fig.width = 8, fig.alt = "Auto and cross regression coefficients plotted against the time interval. The effect of fat on yield grows steadily more negative; the reverse effect stays near zero."}
ctDiscretePars(fit, plot = TRUE)
```

This last plot is the thing continuous time gives you. The same estimated system is
expressed as regression coefficients at every interval between observations, instead
of only at the one interval a discrete time model happens to be fitted at. Over the
ten days shown, the effect of fat on yield keeps accumulating while the reverse
effect stays flat and near zero.

## Covariate effects on parameters

Something closer to real use: chick growth under four diets, asking whether diet
moves a parameter of the process rather than only the observed mean. `ChickWeight`
is in base R.

```{r chickdata}
chickdata <- ChickWeight
# Diet is categorical, so it needs dummy / 'one hot' coding.
chickdata <- cbind(chickdata, model.matrix(~ Diet - 1, chickdata))
colnames(chickdata)[5:8] <- paste0('Diet_', 1:4)
# Scaling continuous variables makes for easier estimation and more sensible
# default priors. Time intervals can benefit too.
chickdata$weight <- scale(chickdata$weight)
head(chickdata, 3)
```

```{r chickmodel, results = 'hide'}
m <- ctModel(
  type = 'ct',
  LAMBDA = diag(1),
  manifestNames = 'weight',
  latentNames = 'Lweight',
  time = 'Time',
  id = 'Chick',
  DRIFT = 'a11, transform=param',              # lift the usual sign restriction
  MANIFESTMEANS = 0,                           # identification: fix one of
  CINT = 'cint, tipreds=c(Diet_2,Diet_3,Diet_4)',   # these two, estimate the other
  TIpredNames = paste0('Diet_', 2:4),          # diet 1 is the baseline
  tipredDefault = FALSE)                       # only the effects asked for

f <- ctFit(chickdata, m, priors = TRUE)
```

Priors are used here because the model is only weakly identified by these data
without them: growth over a dozen observations does not pin down an auto effect and
a continuous intercept independently.

Two pieces of syntax recur everywhere and are worth unpacking:

- **A string is a free parameter, a number is a fixed value.** The observation
  intercept and the process intercept cannot both be free -- they describe the same
  level -- so `MANIFESTMEANS = 0` is what lets `CINT` be estimated.
- **A parameter carries options after its name, as named fields.** The fields are
  `transform`, `indvarying` (random effects), `sdscale` (the random effect sd prior
  relative to the mean's) and `tipreds` (covariate effects); anything not named
  takes its default. So `'cint, tipreds=c(Diet_2,Diet_3,Diet_4)'` is a parameter
  named `cint`, default in every other respect, moderated by three columns.

  The same cell can be written positionally, with `|` separating the fields in that
  order: `'cint ||||Diet_2,Diet_3,Diet_4'` builds an identical model. That is the
  form the manual and older code use, and `ctParSpec()` writes it for you from named
  arguments.

```{r chicktipreds}
summary(f)$tipreds
```

Diets 3 and 4 raise the continuous intercept clearly; diet 2 is not distinguishable
from the baseline diet.

```{r chickpredict, fig.height = 4.2, fig.width = 8, fig.alt = "Predicted weight trajectories for one chick from each of the four diets, based on the diet covariates alone, separating into four increasing curves."}
# One chick per diet, predicted from the covariates alone rather than from data.
ctPredict(f, plot = TRUE, removeObs = TRUE, polygonalpha = 0,
  subjects = as.numeric(chickdata$Chick[!duplicated(ChickWeight$Diet)]))
```

```{r chickdiscrete, fig.height = 4.2, fig.width = 8, fig.alt = "Temporal auto regression coefficient of latent weight plotted against the time interval, rising over the interval range."}
# Temporal regression coefficients as a function of the time interval.
ctDiscretePars(f, plot = TRUE)
```

Because the sign restriction on the auto effect was lifted, the coefficient rises
above one as the interval grows: an increase in weight promotes further increase,
which is what growth looks like in this parameterisation.

## The Julia backend

Alongside the default Stan based fitting, ctsem can fit with a Julia backend. It
is under active development, is generally faster, and supports non-Gaussian
indicators -- binary, ordinal, count and censored -- which the Stan path does not.

```r
ctJuliaInstall()
fit <- ctFit(data, model, backend = 'julia')
```

`ctJuliaInstall()` installs whatever is missing -- Julia itself, the
JuliaConnectoR bridge, and the engine's own dependencies -- and asks before
downloading anything. `ctJuliaStatus()` reports what is present.
[JULIA-BACKEND.md](JULIA-BACKEND.md) covers what the backend supports, where it
improves on Stan rather than reproducing it, and how it is verified against the
Stan path.

## Where to go next

| If you want to | Read |
|---|---|
| Understand what a dynamic system is, and why it is worth the trouble | [Dynamic systems modeling in psychology](https://cdriver.netlify.app/post/introtodynamics/) |
| Work an applied example end to end, from theory through simulation to fitting and checking | the [tutorial preprint](https://osf.io/preprints/psyarxiv/4q9ex_v3), with code and materials in the [OSF project](https://osf.io/xh6ue/) |
| A second, shorter worked start | [ctsem quick start: sunspots and a damped linear oscillator](https://cdriver.netlify.app/post/ctsem-quick-start/) |
| Get everything ctsem can say about a fit, in one folder | `ctReport(fit)` |
| The full reference: the model, the priors, every argument | the [hierarchical manual (pdf)](https://cran.r-project.org/web/packages/ctsem/vignettes/hierarchicalmanual.pdf), or `ctDocs()` in R |
| Understand and control the uncertainty ctsem reports | [uncertainty vignette](https://github.com/cdriveraus/ctsem/blob/juliaFit/vignettes/uncertainty.qmd) |
| Build a measurement model: multiple indicators, non-Gaussian items | [measurement models vignette](https://github.com/cdriveraus/ctsem/blob/juliaFit/vignettes/measurement-models.qmd) |
| Put random effects on parameters by Laplace approximation | [multilevel Laplace vignette](https://github.com/cdriveraus/ctsem/blob/juliaFit/vignettes/laplace-multilevel.qmd) |
| Fit each subject separately, with the population estimate as a prior | [empirical Bayes vignette](https://cran.r-project.org/web/packages/ctsem/vignettes/empirical-bayes-fitting.html) |
| Fit discrete time models, and know when the choice matters | [Discrete time models using ctsem?](https://cdriver.netlify.app/post/dtbivariate/) and [Representations of dynamic systems](https://cdriver.netlify.app/post/discretetime/) |
| Handle an accelerated longitudinal or multiple group design | [Accelerated longitudinal and multiple group designs](https://cdriver.netlify.app/post/accelerated/) |
| Understand missingness, and filtered against smoothed estimates | [Kalman filter vs smoother](https://cdriver.netlify.app/post/missingdata/) |
| Think about individual differences and what heterogeneity means | [What is heterogeneity, really?](https://cdriver.netlify.app/post/heterogeneity/) and the [conceptual paper](https://www.researchgate.net/publication/324093594_Hierarchical_Bayesian_Continuous_Time_Dynamic_Modeling) |
| Model an intervention or another external input | [Understanding the time course of interventions](https://www.researchgate.net/publication/328221807_Understanding_the_Time_Course_of_Interventions_with_Continuous_Time_Dynamic_Models) |
| Fit a growth curve with state dependent error | [Latent growth curves, state dependent error](https://cdriver.netlify.app/post/lgc/) |
| Work with binary observations | [Binary data in state space models](https://cdriver.netlify.app/post/binarydata/) |
| Fit binary, ordinal, count or censored indicators, or want more speed | the [Julia backend](JULIA-BACKEND.md) |
| Point and click instead of writing code | [ctsemGUI](https://github.com/cdriveraus/ctsemgui), a Shiny front end |
| Use the original OpenMx based ctsem (version 1 style) | [ctsemOMX](https://github.com/cdriveraus/ctsemOMX) and its [vignette](https://cran.r-project.org/package=ctsemOMX/vignettes/ctsemOMX.pdf) |

## Function map

The functions most users need, by task. `?name` gives the details for any of them.

**Specify a model**

| | |
|---|---|
| `ctModel` | specify a continuous (`type = 'ct'`) or discrete (`'dt'`) time model |
| `ctModelLatex` | render the model as equations, to pdf or png |
| `model$matrices` | view and edit the specification in matrix form, e.g. `model$matrices$DRIFT[1, 2] <- 'cross'` |
| `ctModelHigherOrder` | raise the order of a model, for instance to allow oscillation |
| `ctModelConvertOMX` | convert an old OpenMx style (`type = 'omx'`) model object |
| `ctParSpec` | write a cell specification from named arguments |
| `ctIdentify` | check which parameters a dataset can inform, before fitting |

**Prepare data**

| | |
|---|---|
| `ctWideToLong`, `ctLongToWide` | reshape between wide and long format |
| `ctIntervalise`, `ctDeintervalise` | convert between absolute times and intervals |
| `ctDiscretiseData` | bin long format observations onto a chosen grid |

**Fit**

| | |
|---|---|
| `ctFit` | fit a model: maximum likelihood, maximum a posteriori, or sampling |
| `ctFitUpdate`, `ctFitUpdateModel` | refit with changed data or specification, without recompiling |
| `ctFitAddSamples` | add posterior samples to an existing fit |
| `ctSample` | sample the posterior of a Julia backend fit |
| `ctEmpiricalBayesFit` | subject-wise fits with the population estimate as a prior |

**Inspect a fit**

| | |
|---|---|
| `summary` | estimates, random effect sds, covariate effects, residual covariance |
| `ctSummaryMatrices` | the dynamic system parameters in matrix form |
| `ctSubjectPars` | subject specific parameters |
| `ctExtract` | the raw draws, as named arrays |
| `ctPlotPosterior` | prior against posterior, per parameter |
| `ctTIpredEffects`, `ctPredictTIP` | covariate effects on parameters, and on expected observations |
| `ctDiscretePars`, `ctDiscreteParsPlot` | temporal regressions as a function of the time interval |
| `ctNetwork`, `ctNetworkPlot` | temporal and contemporaneous networks of the fitted system |
| `ctPhasePortrait` | phase portrait of the fitted dynamics |
| `ctStateDependencePlot` | how model matrix cells vary with a latent state, for nonlinear models |
| `ctTracePlot` | the optimisation trace of a fit |

**Predict**

| | |
|---|---|
| `ctPredict` | predictions and uncertainty, conditional on all, some, or none of the data |
| `ctKalman`, `ctKalmanArray` | filtered and smoothed state estimates |
| `ctResiduals` | standardised residuals |

**Check and compare**

| | |
|---|---|
| `ctReport` | run the reporting and fit-checking functions over a fit and write a folder, with an index saying how to read each file |
| `ctPostPredict` | posterior predictive comparison of implied and observed data |
| `ctFitCheckCov` | lagged covariance diagnostics |
| `ctCoverageCheck` | generate from a model, refit repeatedly, and check parameter recovery |
| `ctParticleLik`, `ctParticleCorrect` | measure the filter's approximation error against a particle filter, and correct the draws for it |
| `ctACF`, `ctACFresiduals` | continuous time autocorrelation, of the data and of the residuals |
| `ctChisqTest` | compare two nested fits |
| `ctLOO` | k fold cross validation |
| `ctLaplaceCheck`, `ctLaplaceCorrect` | measure, and correct, the Laplace approximation error for random effects |

**Generate data**

| | |
|---|---|
| `ctGenerate` | from a model with fixed parameter values |
| `ctGenerateFromPriors` | from a model with free parameters, drawing from the priors |
| `ctGenerateFromFit` | from a fitted model |

**Julia backend**

| | |
|---|---|
| `ctJuliaInstall` | install Julia, the bridge package, and the engine's dependencies |
| `ctJuliaStatus` | report whether the Julia backend is available |
| `ctJuliaSetup` | point ctsem at a particular Julia installation |
| `ctJuliaWorkersStop` | release the warmed sampling worker pool |

## Getting help

Questions, and the answers to a good many earlier ones, are at
<https://github.com/cdriveraus/ctsem/discussions>. Bug reports belong in
[issues](https://github.com/cdriveraus/ctsem/issues).

To cite ctsem, run `citation('ctsem')`.

<details><summary><b>Older function names</b></summary>

From version 3.11.0 the user facing functions dropped their Stan specific wording.
**Existing code keeps working**: the old names remain as aliases.

`ctFit()` is the main fitting function and `ctStanFit()` is an alias of it. Its
second argument is now `model`; the older `ctstanmodel` is deprecated but still
accepted. `ctModelConvertOMX()` replaces `ctStanModel()` for converting old OpenMx
style model objects. Otherwise:

| old | new |
|---|---|
| `ctStanGenerate()` | `ctGenerateFromPriors()` |
| `ctStanGenerateFromFit()` | `ctGenerateFromFit()` |
| `ctStanKalman()` | `ctKalmanArray()` |
| `ctStanPlotPost()` | `ctPlotPosterior()` |
| `ctStanPostPredict()` | `ctPostPredict()` |
| `ctStanSubjectPars()` | `ctSubjectPars()` |
| `ctStanTIpredeffects()` | `ctTIpredEffects()` |
| `ctStanFitUpdate()` | `ctFitUpdate()` |
| `ctStanDiscretePars()` | `ctDiscretePars()` |
| `ctStanDiscreteParsPlot()` | `ctDiscreteParsPlot()` |
| `ctStanContinuousPars()` | `ctSummaryMatrices()` |
| `ctStanParnames()` | `ctRawParnames()` |

</details>

<details><summary><b>Project history</b></summary>

Version 1 of ctsem provided SEM based functionality by linking to OpenMx, which
allowed random means but fixed regression and variance parameters across subjects.
Those functions now live in the separate
[ctsemOMX](https://github.com/cdriveraus/ctsemOMX) package, documented in a JSS
publication (Driver, Voelkle & Oud, 2017) and its
[vignette](https://cran.r-project.org/package=ctsemOMX/vignettes/ctsemOMX.pdf).

Version 2 introduced the hierarchical specification and a Kalman filter coded in
Stan, letting every parameter of the dynamic model vary across individuals with an
estimated population mean and variance. Version 3 added state dependent (time
varying) parameters. For most purposes the current formulation is faster, more
robust and more flexible than the OpenMx one; with many subjects, few time points
and no individual differences in timing, ctsemOMX can still be quicker.

Recent changes are in [NEWS.md](https://github.com/cdriveraus/ctsem/blob/master/NEWS.md).

</details>
