---
title: "Item response models over time"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Item response models over time}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
  fig.height = 4, message = FALSE, warning = FALSE)
```

```{r libs}
library(ctsem)
library(ggplot2)
```

A binary item measured repeatedly is an item response model whose latent
ability moves. ctsem fits the two, three and four parameter logistic forms
with `manifesttype = 1`, and the parameters map onto ones already in the
model:

| IRT | ctsem | notes |
|---|---|---|
| discrimination $a$ | `LAMBDA` | free loading |
| difficulty $b$ | `MANIFESTMEANS` | enters as $-ab$ |
| guessing $c$ | `asymptotes = 1` | lower asymptote |
| upper asymptote $d$ | `asymptotes = 2` | also frees $c$ |

So the two parameter model needs no special argument at all, and `asymptotes`
adds the other two. Ordinal items are in the measurement models vignette; this
one is about binary items and what has to be held fixed to estimate them.

# The response curves

$P(y=1 \mid \eta) = c + (d-c)\,\text{logit}^{-1}(a(\eta - b))$. The three forms
differ only in where the curve starts and stops:

```{r curves, fig.height = 3.4}
eta <- seq(-4, 4, length.out = 300)
curve3 <- function(a, b, c, d, lab) data.frame(eta = eta, lab = lab,
  p = c + (d - c) / (1 + exp(-a * (eta - b))))

ggplot(rbind(
  curve3(1.5, 0, 0,    1,    '2PL'),
  curve3(1.5, 0, 0.25, 1,    '3PL: guessing 0.25'),
  curve3(1.5, 0, 0.25, 0.9,  '4PL: and a ceiling of 0.9')),
  aes(eta, p, colour = lab)) +
  geom_hline(yintercept = c(0, 1), linewidth = .3, colour = 'grey80') +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_bw() + theme(legend.position = 'bottom') +
  labs(x = 'latent ability', y = 'P(correct)', colour = NULL)
```

The guessing parameter lifts the floor: someone far below the item's
difficulty still answers correctly a quarter of the time. The upper asymptote
lowers the ceiling, for items even able people sometimes miss.

# What has to be fixed

The latent process needs a **scale** and a **location**, as in any factor
model, and the logistic link supplies neither. Multiply the process by $k$ and
divide every discrimination by $k$ and the linear predictor is unchanged, so
the likelihood cannot tell them apart.

There are two ways to pin the scale and they are the same model:

* **fix one loading** at 1 and let `DIFFUSION` be free, or
* **fix `DIFFUSION`** at 1 and let every loading be free.

What you cannot do is fix the latent *variance*, because in continuous time it
is not a parameter: the stationary variance is $\text{DIFFUSION}^2/(2|\text{DRIFT}|)$,
so constraining it would mean constraining a function of two things the model
is estimating. Pick one of the two above instead.

For the location, leave `CINT` and `T0MEANS` at zero and let `MANIFESTMEANS`
carry the difficulties. Freeing both ends is the same redundancy again.

`MANIFESTVAR` must be zero: the logistic link already supplies the
randomness, and `ctFit()` fixes it for you if you leave it free.

# Simulating and fitting

Ten items with varying discriminations, 250 people, 8 occasions, guessing at
0.20.

```{r generate}
nitem <- 10
items <- paste0('y', seq_len(nitem))
a <- seq(0.9, 1.9, length.out = nitem)     # discriminations
b <- seq(-1.6, 1.6, length.out = nitem)    # difficulties
guess <- 0.20

gen <- ctModel(type = 'ct', n.latent = 1, n.manifest = nitem,
  manifestNames = items, latentNames = 'eta',
  manifesttype = rep(1, nitem), asymptotes = rep(1, nitem),
  LAMBDA = matrix(a, nitem, 1), MANIFESTMEANS = matrix(-a * b, nitem, 1),
  DRIFT = matrix(-0.4), DIFFUSION = matrix(1.2), T0VAR = matrix(1.342),
  T0MEANS = matrix(0), CINT = matrix(0), MANIFESTVAR = diag(0, nitem),
  Tpoints = 8)

# The guessing probability is THRESHOLDS column 1. Fixing it to generate from
# a known value; in a model being fitted it is left free.
lower <- gen$pars$matrix == 'THRESHOLDS' & gen$pars$col == 1
gen$pars$value[lower] <- guess
gen$pars$param[lower] <- NA
gen$pars$transform[lower] <- NA

set.seed(7)
d <- data.frame(ctGenerate(gen, n.subjects = 250, Tpoints = 8, dtmean = 0.9,
  backend = 'julia'))
```

The floor is visible in the data. An item far above someone's ability is still
answered correctly about `c` of the time, where a 2PL would go to zero:

```{r floor, fig.height = 3.2}
hardest <- d[[items[nitem]]]
ggplot(data.frame(p = c(mean(hardest), guess +
    (1 - guess) * mean(plogis(a[nitem] * (rnorm(1e5, 0, 1.342) - b[nitem]))),
  mean(plogis(a[nitem] * (rnorm(1e5, 0, 1.342) - b[nitem])))),
  lab = factor(c('observed', '3PL expected', '2PL would give'),
    levels = c('observed', '3PL expected', '2PL would give'))),
  aes(lab, p, fill = lab)) +
  geom_col(width = .6, show.legend = FALSE) +
  geom_text(aes(label = sprintf('%.2f', p)), vjust = -0.4) +
  scale_y_continuous(limits = c(0, 0.45)) +
  theme_bw() + labs(x = NULL, y = 'P(correct), hardest item')
```

Fitting needs one line changed between the three models — `asymptotes`.

```{r fit}
irtmodel <- function(asymptotes = NULL) {
  m <- ctModel(type = 'ct', n.latent = 1, n.manifest = nitem,
    manifestNames = items, latentNames = 'eta',
    manifesttype = rep(1, nitem), asymptotes = asymptotes,
    # Scale: the first loading is fixed at 1 and DIFFUSION is free.
    LAMBDA = matrix(c(1, paste0('a_', items[-1])), nitem, 1),
    # Location: the latent is centred, so MANIFESTMEANS are the difficulties.
    CINT = matrix(0), T0MEANS = matrix(0),
    MANIFESTVAR = diag(0, nitem))
  m$pars$indvarying <- FALSE
  # One guessing parameter for the whole test. Per item is possible and much
  # harder to estimate; sharing is the usual starting point.
  if (!is.null(asymptotes)) m$pars$param[m$pars$matrix == 'THRESHOLDS' &
      m$pars$col == 1 & !is.na(m$pars$param)] <- 'guess'
  m
}

fit2 <- ctFit(d[, c('id', 'time', items)], irtmodel(),
  backend = 'julia', cores = 1, priors = TRUE)
fit3 <- ctFit(d[, c('id', 'time', items)], irtmodel(rep(1, nitem)),
  backend = 'julia', cores = 1, priors = TRUE)
```

```{r compare}
data.frame(model = c('2PL', '3PL'),
  loglik = round(c(fit2$estimate$loglik, fit3$estimate$loglik), 1),
  npars = c(length(fit2$estimate$raw), length(fit3$estimate$raw)),
  guess = c(NA, round(summary(fit3)$popmeans['guess', 'mean'], 3)),
  guess_sd = c(NA, round(summary(fit3)$popmeans['guess', 'sd'], 3)))
```

The guessing parameter is recovered with a usable standard error, and the
three parameter model fits better for the one extra parameter.

# Reading the dynamics

The point of fitting these over time is the process, not the items.
`ctDiscretePars()` gives the autoregression of ability across intervals:

```{r dynamics, fig.height = 3.6}
ctDiscretePars(fit3, times = seq(0, 8, .1), plot = TRUE, indices = 'AR',
  cores = 1)
```

# Four parameters, and when to stop

`asymptotes = 2` frees the ceiling as well. It is reported as a gap: the model
estimates how far the upper asymptote sits from the lower one as a proportion
of the room left, so that $0 \le c < d \le 1$ holds without either parameter
having to know the other's value.

```{r fourpl}
m4 <- irtmodel(rep(2, nitem))
m4$pars$param[m4$pars$matrix == 'THRESHOLDS' & m4$pars$col == 2 &
    !is.na(m4$pars$param)] <- 'uppergap'
fit4 <- ctFit(d[, c('id', 'time', items)], m4, backend = 'julia', cores = 1,
  priors = TRUE)
round(summary(fit4)$popmeans[c('guess', 'uppergap'), ], 3)
```

These data were generated with no ceiling, so the four parameter model is
fitting one that is not there — worth seeing, because it is the usual
situation. Asymptotes are weakly informed even in cross-sectional item
response modelling with large samples, and here the dynamics compete for the
same information. Three practical points:

* **Share the asymptote across items** unless you have a reason and the data
  to do otherwise.
* **Use `priors = TRUE`.** The default prior puts the guessing probability
  near 0.18 and the ceiling near 1, which is where they belong when the data
  are quiet about them.
* **Check the fit converged** rather than reading the estimate. A guessing
  parameter reported without a standard error, or alongside a warning about
  curvature, is a stopping point rather than an answer.

# Summary

* A free loading on a binary indicator is already a 2PL; `asymptotes = 1` and
  `2` add the third and fourth parameters.
* Fix either one loading or `DIFFUSION` to set the latent scale. The
  stationary variance cannot be fixed directly, being
  $\text{DIFFUSION}^2/(2|\text{DRIFT}|)$.
* Keep `CINT` and `T0MEANS` at zero so `MANIFESTMEANS` are the difficulties,
  and leave `MANIFESTVAR` at zero.
* Share asymptotes across items and fit with priors; treat an asymptote
  without a standard error as not estimated.
