---
title: "Multilevel random effects by Laplace approximation"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
vignette: >
  %\VignetteIndexEntry{Multilevel random effects by Laplace approximation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.6,
  message = FALSE,
  warning = FALSE
)
```

```{r libs}
library(ctsem)
library(ggplot2)
```

`intoverpop = 'laplace'` integrates individual differences in parameters out of
the likelihood by a Laplace approximation, instead of carrying them as extra
latent states. It accepts a *vector* of id columns, so subjects can be nested in
studies, sites, or classrooms, with random effects declared at each level.

This vignette covers what the approximation is, how to specify and fit such a
model, and how to read, plot and extract from the result. It requires the Julia
backend.

# The model

## Where random effects live

ctsem works internally on a raw, unconstrained parameter vector $\theta$; each
model matrix cell is a function of one raw value through that parameter's
transform. When parameters vary across individuals, subject $i$ gets its own
raw vector rather than its own model:

$$
v_i \;=\; \theta \;+\; \sum_{l} L_l\, u_{b_l(i)}, \qquad u \sim N(0, I)
$$

Here $l$ indexes the levels of the hierarchy, $b_l(i)$ is the group that
subject $i$ belongs to at level $l$ (itself at the subject level, its study at
the study level), and $L_l$ is the Cholesky factor of that level's population
covariance, built from raw standard-deviation and correlation parameters. The
shifts are additive on the raw scale and are applied before the transforms, so
a subject's parameter and the population parameter are always on the same
scale.

Because $u$ is standard normal by construction, everything about the spread of
individual differences lives in $L_l$ — which is what `popsd` and `rawpopcorr`
report.

## Two ways to remove $u$

The likelihood we want is the marginal one, with $u$ integrated out. ctsem has
two ways to do that.

**Augmentation** (`intoverpop = TRUE`) appends $u$ to the latent state with no
drift and no diffusion, so the Kalman filter carries it along and integrates it
as a by-product of filtering. It is exact for a linear-Gaussian model and needs
no new machinery. The cost is that every subject's filtered system grows from
$n$ latent processes to $n + k$, where $k$ is the number of varying parameters,
and the filter's linear algebra is cubic in that dimension.

**Laplace** (`intoverpop = 'laplace'`) leaves every subject's system at $n$ and
approximates the integral instead. For a group of members $U$ that share a
random effect,

$$
T(\theta) \;=\; \log \int \exp\Big( \textstyle\sum_{m \in U} \ell_m(v_m) - \tfrac{1}{2}u'u \Big)\, du
\;-\; \tfrac{d}{2}\log 2\pi
$$

is approximated by expanding the exponent about its mode $\hat u$:

$$
T(\theta) \;\approx\; \sum_{m \in U} \ell_m\big(v_m(\theta, \hat u)\big) \;-\; \tfrac{1}{2}\hat u'\hat u \;-\; \tfrac{1}{2}\log\det M,
\qquad M = I - A
$$

where $A$ is the Hessian of $\sum_m \ell_m$ with respect to $u$ at the mode.
The identity in $M$ is the standard normal prior's own curvature.

The mode is found by an inner Newton solve, warm-started from the previous
evaluation, and the outer gradient with respect to $\theta$ is **exact** — the
approximation is in the integral, not in its derivatives.

## What is exact and what is not

When a random effect enters the state mean linearly through an identity
transform in a linear-Gaussian model, the integrand is exactly Gaussian in $u$,
and Laplace is not an approximation at all — it reproduces the closed-form
integral. `T0MEANS`, `CINT` and `MANIFESTMEANS` are identity-transformed by
default, so a model like the one below is in that regime.

It is genuinely approximate when the effect enters through a nonlinear
transform (`DRIFT`, `DIFFUSION`, and any state-dependent parameter), and the
approximation is better the more informative each subject's data is about its
own effects.

## Nesting

Subjects in the same study share that study's random effect, so their
likelihood contributions do not factorise and they cannot be integrated
separately. The unit of integration is therefore the **outermost group**: one
study, with all of its subjects, is integrated jointly.

That would be expensive done naively — the integral for a study of 100 subjects
with 2 subject effects and 1 study effect is 201-dimensional. It is cheap
because the curvature has an *arrow* structure: each subject's block couples to
its study and to nothing else. ctsem assembles, factorises and inverts that
structure block by block, so the cost is linear in the number of subjects
rather than cubic.

# A worked example

## Simulating

Thirty subjects in five studies, ten occasions each. A univariate
Ornstein-Uhlenbeck process where subjects differ in their starting level
(`T0MEANS`) and in their equilibrium (`CINT`), the two correlated $-0.4$, and
where studies additionally differ in average equilibrium.

```{r simulate}
set.seed(4)
nstudy <- 5; npersub <- 6; nobs <- 10
mu_cint <- 0.3; drift <- -0.4; diffusion <- 0.6; merror <- 0.2
sd_study_cint <- 0.5
sd_sub_cint <- 0.4; sd_sub_t0 <- 0.8; rho <- -0.4

S <- matrix(c(sd_sub_t0^2, rho * sd_sub_t0 * sd_sub_cint,
              rho * sd_sub_t0 * sd_sub_cint, sd_sub_cint^2), 2, 2)
L <- t(chol(S))
studyeff <- rnorm(nstudy, 0, sd_study_cint)

rows <- list(); truth <- list(); id <- 0
for (s in 1:nstudy) for (j in 1:npersub) {
  id <- id + 1
  z <- L %*% rnorm(2)
  t0_i <- as.numeric(z[1])
  cint_i <- mu_cint + studyeff[s] + as.numeric(z[2])
  truth[[id]] <- data.frame(id = id, study = s, t0 = t0_i, cint = cint_i)
  x <- t0_i; y <- numeric(nobs)
  A <- exp(drift); b <- (cint_i / drift) * (A - 1)
  Q <- diffusion^2 / (-2 * drift) * (1 - A^2)
  for (t in 1:nobs) {
    if (t > 1) x <- A * x + b + rnorm(1, 0, sqrt(Q))
    y[t] <- x + rnorm(1, 0, merror)
  }
  rows[[id]] <- data.frame(id = id, study = s, time = 0:(nobs - 1), Y1 = y)
}
dat <- do.call(rbind, rows)
truth <- do.call(rbind, truth)
head(dat)
```

Note that `study` is an ordinary column of the long data frame, constant within
subject. Nesting is read from the data, not declared separately.

## Specifying the hierarchy

Pass `id` a vector, innermost first:

```{r model}
m <- ctModel(type = 'ct', id = c('id', 'study'),
  manifestNames = 'Y1', latentNames = 'eta1', LAMBDA = matrix(1),
  CINT = matrix('cint_eta1'), MANIFESTMEANS = matrix(0))
```

`CINT` is fixed to zero by default in ctsem, with the level carried by
`MANIFESTMEANS`. Here it is freed and `MANIFESTMEANS` fixed instead, so that
the process has a genuine equilibrium at $-\text{CINT}/\text{DRIFT}$ for
individual differences to act on.

Each id gets its own `indvarying` column. The first id keeps the existing
unsuffixed name; every additional one is suffixed with the column name:

```{r columns}
grep('indvarying|sdscale', names(m$pars), value = TRUE)
```

So `indvarying` declares subject-level effects and `indvarying_study`
study-level ones. A parameter may vary at several levels at once — `cint_eta1`
does below, which is what makes its variance decompose into a within-study and
a between-study part:

```{r declare}
m$pars$indvarying <- FALSE
m$pars$indvarying_study <- FALSE
m$pars$indvarying[m$pars$matrix %in% c('T0MEANS', 'CINT')] <- TRUE
m$pars$indvarying_study[m$pars$matrix %in% 'CINT'] <- TRUE

m$pars[!is.na(m$pars$param),
  c('matrix', 'param', 'indvarying', 'indvarying_study')]
```

`sdscale` and `sdscale_study` scale the prior on each level's standard
deviations, defaulting to 1. Levels must be **strictly nested**: every subject
belongs to exactly one study. ctsem checks this and stops if the data say
otherwise.

## Fitting

```{r fit}
fit <- ctFit(dat, m, backend = 'julia', intoverpop = 'laplace', cores = 4)
```

`cores` threads the loop over integration units, which are studies here.
Everything else is as usual: `intoverpop = 'laplace'` requires
`backend = 'julia'` and an optimised (not sampled) fit.

# Reading the output

```{r summary}
summary(fit)
```

The parts specific to a hierarchy:

**`Random-effects standard deviations (id)`** and **`(study)`** are the spread
of the random effects at each level, on the transformed parameter scale. With
more than one level, nothing is reported in an unlabelled block — a bare table
would leave it ambiguous whether it described spread between subjects or
between studies.

Read together they are a variance decomposition. `cint_eta1` appears at both
levels, so its total individual variation splits into a within-study part and a
between-study part, against generating values of $0.4$ and $0.5$:

```{r decomp}
s <- summary(fit)
rbind(
  estimated = c(id = s$popsd.id['cint_eta1', 'mean'],
                study = s$popsd.study['cint_eta1', 'mean']),
  truth = c(id = sd_sub_cint, study = sd_study_cint))
```

**`Random-effects correlations (id)`** are correlations between random effects
*on the raw, untransformed scale* — the correlation ctsem parameterises, not a
correlation of the transformed parameters. For identity-transformed parameters
like these the two coincide; for `DRIFT` or `DIFFUSION` they do not. The study
level has only one effect here, so it has no correlation table.

**`Fixed-effects / Population means`** are the population values, i.e. the
parameters before any subject's shift is added.

For programmatic use, `summary(fit)$randomEffects` holds the same quantities
per level, as raw draws rather than as summaries:

```{r payload}
str(lapply(summary(fit)$randomEffects, function(z)
  list(level = z$level, popsd = colnames(z$popsd))))
```

# Per-subject parameters

`ctSubjectPars()` returns each subject's own parameter values — population
value, plus its random effects at every level it belongs to, plus any
time-independent predictor effects — pushed through the model's transforms:

```{r subjectpars}
sp <- ctSubjectPars(fit)
dim(sp)
round(head(sp[1, , ], 4), 3)
```

The array is `draws x subjects x parameters`. On a Laplace fit there is one
draw: these are the conditional modes at the estimated population parameters, a
point estimate rather than a posterior. Getting draws would mean re-solving
every subject's inner mode at every draw, which is a real computation rather
than a lookup.

Against the values that generated the data:

```{r recovery}
c(cint = cor(sp[1, , 'cint_eta1'], truth$cint),
  T0means = cor(sp[1, , 'T0m_eta1'], truth$t0))
```

Because `cint_eta1` varies at both levels, its subject estimates carry the study
structure, and averaging them within study recovers the study effects:

```{r studymeans}
round(rbind(
  estimated = tapply(sp[1, , 'cint_eta1'], truth$study, mean),
  truth = mu_cint + studyeff), 3)
```

```{r studyplot, fig.height = 4}
d <- data.frame(cint = sp[1, , 'cint_eta1'], study = factor(truth$study))
ggplot(d, aes(x = study, y = cint, colour = study)) +
  geom_hline(yintercept = mean(d$cint), linetype = 'dashed', colour = 'grey50') +
  geom_point(position = position_jitter(width = .12, height = 0), size = 2) +
  stat_summary(fun = mean, geom = 'crossbar', width = .5, colour = 'black') +
  labs(y = 'subject CINT', title = 'Subject equilibria, grouped by study') +
  theme_bw() + theme(legend.position = 'none')
```

The spread of the black study means around the dashed grand mean is the
between-study variation; the scatter within each study is the within-study
variation.

# Trajectories

`ctPredict()` and `ctKalman()` work as they do for any other fit:

```{r predict}
k <- ctPredict(fit, subjects = c(1, 7, 13), timestep = .25,
  timerange = c(0, 9))
plot(k)
```

## Choosing which levels enter the trajectory

`randomEffects` selects which levels the trajectory is built from. Naming an id
column includes that level and every level *outside* it, so:

- `'id'` (the default, the innermost id) gives each subject its own trajectory;
- `'study'` gives the study mean, shared by every subject in a study;
- `'population'` includes none, giving the trajectory implied by the population
  parameters alone.

The effect is clearest in the expectation at each subject's first observation,
which is driven by the parameters rather than by data:

```{r levels}
subs <- c(1, 7, 13, 19, 25)   # one from each study
sapply(c('id', 'study', 'population'), function(lev) {
  p <- ctPredict(fit, subjects = subs, timestep = .5, timerange = c(0, 9),
    randomEffects = lev)
  round(p$value[p$Element %in% 'yprior' & p$Time == 0], 3)
})
```

Under `'id'` each subject starts at its own estimated `T0MEANS`. Under `'study'`
and `'population'` they all start at the same place, because `T0MEANS` varies
only at the subject level here — the study effect is on `CINT`, so it acts on
where the trajectories go, not on where they begin.

## The smoothing caveat

A Laplace fit warns, once per call, that its trajectories are the *smoothed*
equivalent. This is worth understanding rather than dismissing.

In an augmented fit the random effects are latent states, so the filter updates
them observation by observation, and a trajectory at time $t$ can be built from
only the data up to $t$. Under Laplace they are conditional modes estimated from
each subject's whole record. There is no version of them from before a
subject's later data arrived, so even `yprior` is conditional on the complete
series. Comparisons that depend on genuine out-of-sample filtering — a
prediction at $t$ that has not seen $t$ — are not what these give.

Relatedly, `removeObs` withholds observations from the filter, and on a Laplace
fit that also removes what the modes are estimated from: with no data left to
condition on, the modes collapse to zero and the trajectories fall back to the
population parameters whatever `randomEffects` asks for.

# Practical notes

**How many groups at the outer level.** A level's standard deviation is
estimated from the number of *groups* at that level, not the number of
subjects. Five studies is few, and the interval on `popsd.study` above is
correspondingly wide. With very few groups the correlation parameters can be
driven to the boundary; ctsem caps them just short of $\pm 1$ and reports that
it has done so rather than failing.

**Priors** are available at every level and are the standard remedy when an
outer level is weakly identified.

**When to prefer augmentation.** Augmentation is faster on small models and is
exact for linear-Gaussian ones, so it remains the default. Laplace is worth
reaching for when the augmented state becomes the bottleneck — many varying
parameters, so that $n + k$ is much larger than $n$ — and it is the only option
for nesting beyond the subject level.

**Time-independent predictors** work at any level, provided the predictor
actually varies at that level.

# Session

```{r session}
sessionInfo()
```
