---
title: "Measurement models"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Measurement models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.4,
  message = FALSE,
  warning = FALSE
)
```

```{r libs}
library(ctsem)
library(ggplot2)
```

A ctsem model has two halves. The **dynamic** half says how the latent
processes evolve between observations; the **measurement** half says how an
observation at a time point relates to the latent state then. This vignette is
entirely about the second half: how many indicators load on a process, what
kind of variable each indicator is, what happens when kinds are mixed, and what
to do when a loading or a residual is not a constant.

Everything here uses `backend = 'julia'`. The non-Gaussian indicator types
(ordinal, count, censored) exist only on that backend, and binary is
*integrated* there rather than linearised, so a vignette that switched backends
between sections would not be comparing like with like.

# The measurement equation

At each observed time point the filter has a predicted latent state $\eta$. For
every manifest variable $j$ observed at that row, ctsem forms a linear
predictor

$$
\mu_j \;=\; \text{MANIFESTMEANS}_j \;+\; \text{LAMBDA}_{j\cdot}\, \eta
$$

and then attaches an observation model to it, chosen by that variable's
`manifesttype`:

| `manifesttype` | Variable | Observation model | Residual |
|---|---|---|---|
| `0` | continuous | $y_j \sim N(\mu_j,\ \text{MANIFESTVAR})$ | free |
| `1` | binary, coded 0/1 | $P(y_j = 1) = \text{logit}^{-1}(\mu_j)$ | none |
| `2` | ordinal, coded $1..K$ | $P(y_j \le k) = \text{logit}^{-1}(\tau_{jk} - \mu_j)$ | none |
| `3` | count, non-negative integers | $y_j \sim \text{Poisson}(e^{\mu_j})$ | none |
| `4` | censored | $N(\mu_j,\ \text{MANIFESTVAR})$, observed only inside $[\,c^-_j, c^+_j\,]$ | free |

Three things follow from that table and are worth stating before any code.

**Only types 0 and 4 have a residual variance.** For binary, ordinal and count
indicators the randomness is entirely in the link — a Bernoulli draw at a given
probability has no separate error term to estimate. Their rows and columns of
`MANIFESTVAR` must be fixed to zero.

**The scale of the latent process has to come from somewhere.** With a Gaussian
indicator you can anchor it by fixing a loading, or by fixing `DIFFUSION`. With
only binary or ordinal indicators there is no free residual to trade against,
so a loading and the process scale are two names for one quantity: fix the
loadings and let `DIFFUSION` carry the scale.

**Location works the same way.** `CINT` and `T0MEANS` set where the process
sits, and each indicator's `MANIFESTMEANS` sets where that indicator sits
relative to it. ctsem's default fixes `CINT` and `T0MEANS` to zero and frees
`MANIFESTMEANS`, which is what the examples below use. An ordinal indicator is
the exception: its thresholds already carry its location, so its
`MANIFESTMEANS` has to be fixed to zero or the two are not separately
identified. ctsem warns when both are free.

# One indicator or many

A univariate Ornstein-Uhlenbeck process, measured three times over at each
occasion with loadings 1, 0.8 and 1.2:

```{r sim-multi}
set.seed(3)
genmulti <- ctModel(type = 'ct', n.latent = 1, n.manifest = 3,
  manifestNames = c('y1', 'y2', 'y3'), latentNames = 'eta1',
  LAMBDA = matrix(c(1, 0.8, 1.2), 3, 1),
  DRIFT = matrix(-0.4), DIFFUSION = matrix(1.5), T0VAR = matrix(1.7),
  T0MEANS = matrix(0), CINT = matrix(0),
  MANIFESTMEANS = matrix(c(0, 0.5, -0.5), 3, 1),
  MANIFESTVAR = diag(0.8, 3), Tpoints = 10)

dmulti <- data.frame(ctGenerate(genmulti, n.subjects = 50, Tpoints = 10,
  backend = 'julia'))
head(dmulti, 3)
```

## A single indicator

With one indicator per process the loading is not estimable and is fixed, by
convention to 1. That does *not* mean measurement error is inestimable: in a
continuous time model the measurement error and the process innovation have
different signatures across the time interval, so `MANIFESTVAR` and `DIFFUSION`
remain separately identified.

```{r fit-single}
msingle <- ctModel(type = 'ct', n.latent = 1, n.manifest = 1,
  manifestNames = 'y1', latentNames = 'eta1', LAMBDA = matrix(1),
  CINT = matrix(0), T0MEANS = matrix(0))
msingle$pars$indvarying <- FALSE

fsingle <- ctFit(dmulti, msingle, backend = 'julia', cores = 2)
summary(fsingle)$popmeans
```

## Three indicators

Freeing the two remaining loadings and keeping the first at 1 makes `eta1` a
common factor. The first indicator's scale is now the latent's scale, which is
what the fixed 1 buys:

```{r fit-multi}
mmulti <- ctModel(type = 'ct', n.latent = 1, n.manifest = 3,
  manifestNames = c('y1', 'y2', 'y3'), latentNames = 'eta1',
  LAMBDA = matrix(c(1, 'lambda_y2', 'lambda_y3'), 3, 1),
  CINT = matrix(0), T0MEANS = matrix(0))
mmulti$pars$indvarying <- FALSE

fmulti <- ctFit(dmulti, mmulti, backend = 'julia', cores = 2)
summary(fmulti)$popmeans
```

The loadings come back near their generating values, and the dynamic parameters
are estimated more sharply than from one indicator alone:

```{r compare-indicators}
s1 <- summary(fsingle)$popmeans
s3 <- summary(fmulti)$popmeans
pars <- c('drift_eta1', 'diff_eta1')
data.frame(
  parameter = pars,
  truth = c(-0.4, 1.5),
  single = s1[pars, 'mean'], single_se = s1[pars, 'sd'],
  three = s3[pars, 'mean'], three_se = s3[pars, 'sd'],
  row.names = NULL)
```

::: {.callout-note}
## Anchoring, and where the optimiser starts

Fixing `LAMBDA[1,1]` is not the only way to set the scale — fixing `DIFFUSION`
to 1 and freeing all three loadings fits the same model, and reports loadings
on the process's own scale instead of on `y1`'s.

The two are equivalent in principle but not always in practice, because where
the optimiser starts matters. `ctFit` reads the starting values for the
variance diagonals off the data by default, through
`optimcontrol$datastart`. Before that it began every fit from a fixed point
that put `DIFFUSION` near 7 whatever the data were measured in, and on data at
a much smaller scale a loading-anchored model would collapse the latent to
zero variance and explain the whole signal as measurement error — a local
optimum, with the loadings running off to implausible magnitudes to
compensate. On the data above rescaled by 1/100 that failed on four starting
seeds out of four, and converges on four out of four now.

It is not a guarantee. A model can sit near a basin boundary at a given scale,
where the small random jitter ctsem still adds to the start is enough to decide
which optimum the fit reaches — on the same data rescaled by 100, two seeds in
four land 328 log units apart, with or without the derived start. So a fit
reporting non-convergence, or loadings nowhere near anything plausible, is
worth re-running, and worth checking against `DIFFUSION`-anchored scaling or
explicit `inits`.
:::

# The five indicator types

To see each type on its own, generate one latent trajectory and then measure it
five different ways. Generating the process with a near-zero measurement
variance makes the single manifest variable a direct readout of the latent
state, which the indicators below are built from:

```{r sim-latent}
set.seed(5)
nsub <- 80; nobs <- 10
process <- ctModel(type = 'ct', n.latent = 1, n.manifest = 1,
  manifestNames = 'eta', latentNames = 'eta1', LAMBDA = matrix(1),
  DRIFT = matrix(-0.4), DIFFUSION = matrix(1), T0VAR = matrix(1.12),
  T0MEANS = matrix(0), CINT = matrix(0), MANIFESTMEANS = matrix(0),
  MANIFESTVAR = matrix(1e-6), Tpoints = nobs)

d <- data.frame(ctGenerate(process, n.subjects = nsub, Tpoints = nobs,
  backend = 'r'))
eta <- d$eta
```

Two indicators of each kind, built by applying each row of the table above by
hand, so that the link in the model and the link in the simulation are visibly
the same thing:

```{r sim-types}
invlogit <- function(x) 1 / (1 + exp(-x))
ordinaldraw <- function(mu, tau) {          # cumulative logit
  cum <- sapply(tau, function(k) invlogit(k - mu))
  p <- cbind(cum, 1)
  p <- cbind(p[, 1, drop = FALSE], t(apply(p, 1, diff)))
  apply(p, 1, function(pr) sample.int(length(pr), 1, prob = pmax(pr, 0)))
}
tau <- c(-1, 0, 1)                          # four categories

d$y1 <- 0.0 + eta + rnorm(nrow(d), 0, 0.6)              # 0 continuous
d$y2 <- 0.5 + eta + rnorm(nrow(d), 0, 0.6)
d$b1 <- rbinom(nrow(d), 1, invlogit(-0.4 + eta))        # 1 binary
d$b2 <- rbinom(nrow(d), 1, invlogit( 0.4 + eta))
d$o1 <- ordinaldraw(eta, tau)                           # 2 ordinal
d$o2 <- ordinaldraw(eta, tau)
d$c1 <- rpois(nrow(d), exp(1.4 + 0.5 * eta))            # 3 count
d$c2 <- rpois(nrow(d), exp(1.2 + 0.5 * eta))
d$x1 <- pmin(pmax(2.0 + eta + rnorm(nrow(d), 0, 0.5), 0), 4)   # 4 censored
d$x2 <- pmin(pmax(1.8 + eta + rnorm(nrow(d), 0, 0.5), 0), 4)

d$eta <- NULL
head(d, 3)
```

One subject, seen through all five instruments at once:

```{r plot-types, fig.height = 5.2}
shown <- c('y1', 'b1', 'o1', 'c1', 'x1')
long <- do.call(rbind, lapply(shown, function(v)
  data.frame(time = d$time[d$id == 2], value = d[[v]][d$id == 2],
    indicator = v)))
long$indicator <- factor(long$indicator, levels = shown,
  labels = c('y1 continuous', 'b1 binary', 'o1 ordinal', 'c1 count',
    'x1 censored [0,4]'))

ggplot(long, aes(time, value)) +
  geom_line(colour = 'grey60') + geom_point(size = 2) +
  facet_wrap(~indicator, scales = 'free_y', ncol = 2) +
  theme_bw() + labs(x = 'time', y = NULL)
```

## Specifying each type

`manifesttype` is a vector with one entry per manifest variable. Two of the
types need something extra alongside it, and both are needed because
`ctModel()` never sees the data:

* **ordinal** needs `ncategories`, so that the right number of thresholds can
  be created. `ctGenerate()` needs it for the same reason.
* **censored** needs `censormin` and `censormax`. These are known constants
  rather than parameters: an instrument's floor and ceiling are properties of
  the instrument, and the data cannot inform them. A variable censored on one
  side only leaves the other infinite.

A helper, since the five models below differ only in which columns they take:

```{r model-builder}
buildmodel <- function(names, type, lambda = 1, ncat = 0,
  cmin = -Inf, cmax = Inf) {
  n <- length(names)
  # only continuous and censored indicators have a residual to estimate
  mvar <- diag(0, n)
  if (type %in% c(0, 4)) diag(mvar) <- paste0('mvar_', names)
  # an ordinal indicator's location is carried by its thresholds
  mmeans <- if (type == 2) matrix(0, n, 1) else
    matrix(paste0('mm_', names), n, 1)
  m <- ctModel(type = 'ct', n.latent = 1, n.manifest = n,
    manifestNames = names, latentNames = 'eta1',
    manifesttype = rep(type, n), ncategories = rep(ncat, n),
    censormin = rep(cmin, n), censormax = rep(cmax, n),
    LAMBDA = matrix(lambda, n, 1), MANIFESTMEANS = mmeans, MANIFESTVAR = mvar,
    CINT = matrix(0), T0MEANS = matrix(0))
  m$pars$indvarying <- FALSE
  m
}

models <- list(
  continuous = buildmodel(c('y1', 'y2'), 0),
  binary     = buildmodel(c('b1', 'b2'), 1),
  ordinal    = buildmodel(c('o1', 'o2'), 2, ncat = 4),
  count      = buildmodel(c('c1', 'c2'), 3, lambda = 0.5),
  censored   = buildmodel(c('x1', 'x2'), 4, cmin = 0, cmax = 4))
```

Every one of these fixes its loadings rather than freeing them. For binary,
ordinal and count that is forced: with no residual variance to trade a loading
against, a free loading and `DIFFUSION` would be the same parameter twice.

The count pair is fixed at 0.5 rather than 1, and that is a choice worth making
deliberately. A count's loading sets how far its *log* rate travels as the
process moves, and the exponential turns a wide latent range into a very
skewed observation. The filter carries a Gaussian summary of the state, and a
Poisson update is well summarised that way only while the rate stays in a
moderate range. Measured on this design at 200 subjects and 12 occasions: a
count read directly off a process with a standard deviation of 1.12 -- log
rates spanning about seven units, rates from 0.05 to 50 -- recovers `DRIFT` at
-0.20 and -0.32 on two seeds against a generating -0.4, and `DIFFUSION` at
0.69 and 0.74 against 1. The same process at a standard deviation of 0.45
recovers -0.407 and -0.418, and its `DIFFUSION` to within 0.012. Halving the
loading halves the range the link sees, which is why the count column below is
as accurate as the others. If your counts genuinely do swing over orders of
magnitude, treat the dynamic estimates from them with the same suspicion.

## The same process, through five instruments

```{r fit-types}
fits <- lapply(models, function(m) ctFit(d, m, backend = 'julia', cores = 2))

pull <- function(f, p, what) summary(f)$popmeans[p, what]
data.frame(
  type = names(fits),
  drift = sapply(fits, pull, 'drift_eta1', 'mean'),
  drift_se = sapply(fits, pull, 'drift_eta1', 'sd'),
  diffusion = sapply(fits, pull, 'diff_eta1', 'mean'),
  diffusion_se = sapply(fits, pull, 'diff_eta1', 'sd'),
  row.names = NULL)
```

Every row of that table estimates the same two generating values, $-0.4$ and
$1$, from the same latent trajectory, and every row is within about one
standard error of them. What differs is how much of the trajectory each
instrument transmits, which is visible in the standard errors rather than in
the estimates: two binary items per occasion carry a few bits about a
continuous state and are the least precise by a factor of two, while the
continuous, censored and count pairs each recover `DRIFT` to about 0.05.

The parameters specific to a type appear in the same summary. Ordinal
thresholds are reported as the first threshold followed by the **gap** to each
subsequent one, which is what keeps them ordered — cumulate them to read the
thresholds themselves:

```{r thresholds}
thr <- summary(fits$ordinal)$popmeans
thr[grep('threshold_o1', rownames(thr)), c('mean', 'sd')]
cumsum(thr[grep('threshold_o1', rownames(thr)), 'mean'])   # against c(-1, 0, 1)
```

Censoring is the one non-Gaussian type that keeps a residual standard
deviation. The model recovers it from the uncensored part of the scale, while
treating a value sitting at a limit as the probability of being there rather
than as a density:

```{r censored-pars}
summary(fits$censored)$popmeans[c('mvar_x1', 'mvar_x2', 'mm_x1', 'mm_x2'),
  c('mean', 'sd')]
mean(d$x1 %in% c(0, 4))     # proportion of x1 observations at a limit
```

# Combining types

## Several types on one latent

Nothing stops indicators of different types from loading on the same process.
Each contributes its own likelihood term against the same predicted state, and
the filter's update at that row takes all of them:

```{r fit-mixed-onelatent}
mixednames <- c('y1', 'b1', 'o1', 'c1', 'x1')
mvar <- diag(0, 5)
mvar[1, 1] <- 'mvar_y1'
mvar[5, 5] <- 'mvar_x1'

mmixed <- ctModel(type = 'ct', n.latent = 1, n.manifest = 5,
  manifestNames = mixednames, latentNames = 'eta1',
  manifesttype = c(0, 1, 2, 3, 4),
  ncategories   = c(0, 0, 4, 0, 0),
  censormin     = c(-Inf, -Inf, -Inf, -Inf, 0),
  censormax     = c( Inf,  Inf,  Inf,  Inf, 4),
  LAMBDA = matrix(c(1, 1, 1, 0.5, 1), 5, 1),   # the count as generated
  MANIFESTMEANS = matrix(c('mm_y1', 'mm_b1', 0, 'mm_c1', 'mm_x1'), 5, 1),
  MANIFESTVAR = mvar, CINT = matrix(0), T0MEANS = matrix(0))
mmixed$pars$indvarying <- FALSE

fmixed <- ctFit(d, mmixed, backend = 'julia', cores = 2)
summary(fmixed)$popmeans
```

One indicator of each kind recovers the process about as precisely as the best
same-kind pair above, and roughly twice as precisely as the binary or ordinal
pairs -- with the thresholds, the two residual standard deviations and four
intercepts all estimated alongside. That is the practical argument for using
indicators of every kind you have, rather than keeping only whichever are
convenient to model.

## Several types on different latents

More usually the types are not interchangeable measures of one thing. An
ordinal questionnaire item and a binary event indicator measure one construct;
a count, a censored scale score and a plain continuous variable measure
another; and the question is how the two constructs drive each other. That is a
two-process model with a block-structured `LAMBDA`.

```{r sim-twolatent}
set.seed(11)
twonames <- c('o1', 'b1', 'c1', 'x1', 'y1')
lambda2 <- matrix(0, 5, 2)
lambda2[1:2, 1] <- 1     # ordinal and binary measure eta1
lambda2[3:5, 2] <- 1     # count, censored and continuous measure eta2

genvar <- diag(0, 5)
genvar[4, 4] <- 0.4
genvar[5, 5] <- 0.6
gen2 <- ctModel(type = 'ct', n.latent = 2, n.manifest = 5,
  manifestNames = twonames, latentNames = c('eta1', 'eta2'),
  manifesttype = c(2, 1, 3, 4, 0), ncategories = c(4, 0, 0, 0, 0),
  censormin = c(-Inf, -Inf, -Inf, 0, -Inf),
  censormax = c(Inf, Inf, Inf, 4, Inf),
  LAMBDA = lambda2,
  DRIFT = matrix(c(-0.4, 0.25, -0.15, -0.7), 2, 2, byrow = TRUE),
  DIFFUSION = diag(0.8, 2), T0VAR = diag(1, 2), T0MEANS = matrix(0, 2, 1),
  CINT = matrix(0, 2, 1),
  MANIFESTMEANS = matrix(c(0, -0.4, 0.5, 2.0, 0.3), 5, 1),
  MANIFESTVAR = genvar, Tpoints = 10)

# THRESHOLDS holds the first threshold, then the gap to each subsequent one
gen2$pars$value[gen2$pars$matrix %in% 'THRESHOLDS' & gen2$pars$col == 1] <- -1
gen2$pars$value[gen2$pars$matrix %in% 'THRESHOLDS' & gen2$pars$col > 1] <- 1

d2 <- data.frame(ctGenerate(gen2, n.subjects = 60, Tpoints = 10,
  backend = 'julia'))
head(d2, 3)
```

`ctGenerate(backend = 'julia')` draws each row from its own prior predictive as
the filter reaches it, so it applies the links above rather than producing
continuous values for a variable the model calls binary. The R generator
integrates a linear Gaussian system and has no notion of a link; it warns if
asked to generate from a model with non-Gaussian indicators.

The estimation model has the same block structure, with the measurement
parameters freed:

```{r fit-twolatent}
estvar <- diag(0, 5)
estvar[4, 4] <- 'mvar_x1'
estvar[5, 5] <- 'mvar_y1'

m2 <- ctModel(type = 'ct', n.latent = 2, n.manifest = 5,
  manifestNames = twonames, latentNames = c('eta1', 'eta2'),
  manifesttype = c(2, 1, 3, 4, 0), ncategories = c(4, 0, 0, 0, 0),
  censormin = c(-Inf, -Inf, -Inf, 0, -Inf),
  censormax = c(Inf, Inf, Inf, 4, Inf),
  LAMBDA = lambda2,
  MANIFESTMEANS = matrix(c(0, 'mm_b1', 'mm_c1', 'mm_x1', 'mm_y1'), 5, 1),
  MANIFESTVAR = estvar, CINT = matrix(0, 2, 1), T0MEANS = matrix(0, 2, 1))
m2$pars$indvarying <- FALSE

f2 <- ctFit(d2, m2, backend = 'julia', cores = 2)
summary(f2)$popmeans
```

The cross-effects are the point of such a model, and they are estimated from
indicators of four different kinds at once:

```{r crosseffects}
p2 <- summary(f2)$popmeans
drifts <- c('drift_eta1', 'drift_eta1_eta2', 'drift_eta2_eta1', 'drift_eta2')
data.frame(
  effect = c('eta1 on itself', 'eta2 on eta1', 'eta1 on eta2', 'eta2 on itself'),
  truth = c(-0.4, 0.25, -0.15, -0.7),
  estimate = p2[drifts, 'mean'],
  se = p2[drifts, 'sd'],
  row.names = NULL)
```

`ctKalman()` reconstructs the latent trajectories from the mixed indicators as
it would from any others:

```{r kalman-twolatent, fig.height = 4.6}
plot(ctKalman(f2, subjects = 1:2, timestep = .2), kalmanvec = 'etasmooth')
```

# State-dependent measurement

A `LAMBDA` or `MANIFESTVAR` cell need not be a constant. Writing a latent
process name into the cell makes it an expression evaluated at the current
state, so an instrument can become more or less sensitive, or noisier, as the
process moves.

Free parameters used inside such an expression have to live in the `PARS`
matrix: a cell containing an expression is the specification itself rather than
a labelled parameter, so ctsem resolves names appearing in it against `PARS`.
`PARS` entries are identity-transformed and unbounded, which is why the
residual below is written as `exp(...)` — that is what keeps a standard
deviation positive.

Data where the second indicator's loading rises with the process, and the first
indicator's residual standard deviation rises with it too:

```{r sim-statedep}
set.seed(9)
gensd <- ctModel(type = 'ct', n.latent = 1, n.manifest = 2,
  manifestNames = c('y1', 'y2'), latentNames = 'eta1',
  LAMBDA = matrix(c(1, '0.9 + 0.35 * eta1'), 2, 1),
  MANIFESTVAR = matrix(c('exp(-0.7 + 0.3 * eta1)', 0,
    0, 0.8), 2, 2),
  DRIFT = matrix(-0.4), DIFFUSION = matrix(1.5), T0VAR = matrix(1.7),
  T0MEANS = matrix(0), CINT = matrix(0),
  MANIFESTMEANS = matrix(c(0, 0.5), 2, 1), Tpoints = 10)

dsd <- data.frame(ctGenerate(gensd, n.subjects = 60, Tpoints = 10,
  backend = 'julia'))
head(dsd, 3)
```

The estimation model writes the same two expressions, with parameter names in
place of the generating constants:

```{r fit-statedep}
msd <- ctModel(type = 'ct', n.latent = 1, n.manifest = 2,
  manifestNames = c('y1', 'y2'), latentNames = 'eta1',
  LAMBDA = matrix(c(1, 'lam0 + lamslope * eta1'), 2, 1),
  MANIFESTVAR = matrix(c('exp(logsd0 + logsdslope * eta1)', 0,
    0, 'mvar_y2'), 2, 2),
  MANIFESTMEANS = matrix(c(0, 'mm_y2'), 2, 1),
  CINT = matrix(0), T0MEANS = matrix(0),
  PARS = matrix(c('lam0', 'lamslope', 'logsd0', 'logsdslope'), 4, 1))
msd$pars$indvarying <- FALSE

fsd <- ctFit(dsd, msd, backend = 'julia', cores = 2)
summary(fsd)$popmeans
```

Against the generating values:

```{r statedep-recovery}
psd <- summary(fsd)$popmeans
sdpars <- c('lam0', 'lamslope', 'logsd0', 'logsdslope')
data.frame(
  parameter = sdpars,
  meaning = c('loading on y2 at eta1 = 0',
    'change in that loading per unit of eta1',
    'log residual sd of y1 at eta1 = 0',
    'change in that log sd per unit of eta1'),
  truth = c(0.9, 0.35, -0.7, 0.3),
  estimate = psd[sdpars, 'mean'],
  row.names = NULL)
```

`ctModelIsNonlinear()` reports that such a model has cells with no single
value, so every summary of it is a reading at one particular state:

```{r nonlinear-flag}
ctModelIsNonlinear(fsd)
```

Which is why the honest picture of a state-dependent model is a curve rather
than a number. `ctStateDependencePlot()` finds the cells that actually move —
by materialising every model matrix across a grid of states and keeping the
ones whose value changes, so dependence arriving indirectly, through a `PARS`
entry or through a covariance built from a dependent cell, is caught too — and
draws each of them against the process:

```{r statedep-plot, fig.height = 4}
ctStateDependencePlot(fsd, along = 'eta1', nsamples = 30)
```

One panel is the loading on `y2`, rising with `eta1`. The other is the residual
*variance* of `y1` — `MANIFESTcov` rather than `MANIFESTVAR`, since that is the
covariance the filter actually uses — rising with it as well. A `summary()` of
this model reports one point from each of those curves.

Only `LAMBDA` and `MANIFESTVAR` are state-dependent here, but the same
mechanism applies to `MANIFESTMEANS`, and on the dynamic side to `DRIFT`,
`DIFFUSION` and `CINT`.

# Summary

* `manifesttype` picks the observation model per indicator: `0` continuous,
  `1` binary, `2` ordinal (with `ncategories`), `3` count, `4` censored (with
  `censormin` and `censormax`). Everything but `0` needs `backend = 'julia'`.
* Only continuous and censored indicators have a residual variance to estimate;
  fix the rest of `MANIFESTVAR` to zero.
* Fix an ordinal indicator's `MANIFESTMEANS` to zero — its thresholds carry its
  location. `THRESHOLDS` holds the first threshold and then the gaps.
* Anchor the latent scale exactly once: a fixed loading, or a fixed
  `DIFFUSION`. With no Gaussian indicator on a process it has to be the
  loadings.
* A count indicator's loading sets how far its log rate travels. Keep that
  range moderate -- the filter's Gaussian summary of a Poisson update degrades
  once the rate swings over orders of magnitude, and the dynamic parameters
  come back attenuated.
* Types can be mixed freely, on one process or across several.
* A `LAMBDA` or `MANIFESTVAR` cell written as an expression in a latent process
  name is evaluated at the current state; free parameters inside it go in
  `PARS`, and `ctStateDependencePlot()` shows what the result claims.
