---
title: "Uncertainty in a Fitted ctsem Model"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Uncertainty in a Fitted ctsem Model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.4,
  message = FALSE,
  warning = FALSE
)
```

```{r libs}
library(ctsem)
library(ggplot2)
set.seed(12)
```

Every `ctFit()` returns an estimate and a claim about how uncertain it is. Where
that claim comes from depends on three choices, each of which changes what the
resulting interval means:

- **`optimize`**: sample the posterior (or likelihood) with HMC, or find its
  mode and approximate the curvature there.
- **`priors`**: include ctsem's weakly informative priors on the raw parameter
  scale, or leave them out.
- **`uncertainty`**: for an optimized fit, which curvature-or-resampling
  approximation to use around the mode.

This vignette works through all three, on both backends (`backend = 'stan'`
and `backend = 'julia'`), using one small model fit many different ways so the
intervals can be compared directly rather than described in the abstract.

## The two routes

**Sampling** (`optimize = FALSE`) draws from the target density with HMC --
Stan's sampler for `backend = 'stan'`, the julia engine's own No-U-Turn
sampler for `backend = 'julia'`. Uncertainty is then just the spread of the
draws: no curvature is computed and no normal approximation is assumed. This
is the most trustworthy route and the slowest one.

**Optimizing** (`optimize = TRUE`, the default) finds the mode of the same
target and reports a point estimate plus an approximation of the local
curvature there. The approximation is cheap -- one fit instead of thousands of
gradient evaluations -- but it is only as good as the normal approximation it
rests on. It is a poor description of a posterior that is skewed, bounded, or
multimodal, which variance-like parameters often are with few subjects.

Both routes target the *same* density for a given setting of `priors`; they
differ only in how they characterise it. The rest of this vignette compares
them directly, then works through what sits between "one Hessian" and "full
HMC," and then through what changes when priors are switched off.

## A small model, fit several ways

The model below is a single latent process observed with a fixed measurement
scale, estimating a continuous-time intercept, drift, and diffusion. Ten
subjects and eight time points keep every fit in this document fast, which
also means several of the demonstrations below hit the small-sample problems
this vignette is about -- deliberately, since a model too well identified to
show the difference between uncertainty methods would not make the point.

```{r generate-data}
n_subjects <- 8
tpoints <- 8

gen_model <- ctModel(
  type = "omx",
  Tpoints = tpoints,
  n.latent = 1,
  n.manifest = 1,
  LAMBDA = matrix(1),
  T0MEANS = matrix(0),
  T0VAR = matrix(0.6),
  DRIFT = matrix(-0.35),
  DIFFUSION = matrix(0.5),
  CINT = matrix(0.15),
  MANIFESTMEANS = matrix(0),
  MANIFESTVAR = matrix(0.25)
)

dat <- do.call(rbind, lapply(seq_len(n_subjects), function(id) {
  d <- data.frame(ctGenerate(
    ctmodelobj = gen_model,
    n.subjects = 1,
    burnin = 0,
    dtmean = 0.4,
    logdtsd = 0.1
  ))
  d$id <- id
  d
}))
dat <- dat[order(dat$id, dat$time), ]

fit_model <- ctModel(
  type = "ct",
  Tpoints = tpoints,
  n.latent = 1,
  n.manifest = 1,
  LAMBDA = matrix(1),
  MANIFESTMEANS = matrix(0),
  CINT = matrix("cint"),
  DRIFT = matrix("drift"),
  DIFFUSION = matrix("diffusion")
)
# A fixed-effects model: no individual differences to estimate on top of the
# already small sample. ctModel() defaults T0MEANS and CINT to individually
# varying, which is realistic for most applications but adds population
# standard deviations and correlations this vignette does not need, and with
# eight subjects those are barely identified on their own -- confounding
# uncertainty about the individual-differences distribution with uncertainty
# about the drift this vignette is actually comparing.
fit_model$pars$indvarying <- FALSE
```

The two backends store an optimized fit's point estimate and posterior draws
in different places (`fit$stanfit$rawest` / `fit$stanfit$rawposterior` for
`backend = 'stan'`, `fit$estimate$raw` / `fit$estimate$rawposterior` for
`backend = 'julia'`). `summary()` and `ctExtract()` read either transparently
and report on the usual constrained parameter scale, so the rest of this
vignette uses those rather than the raw slots directly.

```{r popmeans-helper}
drift_interval <- function(fit) {
  draws <- ctExtract(fit)$pop_DRIFT[, 1, 1]
  c(mean = mean(draws), quantile(draws, c(.025, .975)))
}
```

### Sampling

```{r sample-stan}
stan_sample <- ctFit(
  datalong = dat, model = fit_model, backend = "stan",
  cores = 1, priors = TRUE, optimize = FALSE,
  chains = 1, iter = 400, verbose = 0
)
```

```{r sample-julia}
julia_sample <- ctFit(
  datalong = dat, model = fit_model, backend = "julia",
  cores = 1, priors = TRUE, optimize = FALSE,
  chains = 1, iter = 250, verbose = 0
)
```

With a single short chain each, these are quick demonstrations rather than
publication-ready samples -- raise `chains` and `iter` for real work. The
drift interval from each:

```{r sample-compare}
rbind(stan = drift_interval(stan_sample), julia = drift_interval(julia_sample))
```

### Optimizing

```{r optimize-stan}
stan_map <- ctFit(
  datalong = dat, model = fit_model, backend = "stan",
  cores = 1, priors = TRUE, optimize = TRUE,
  optimcontrol = list(stochastic = FALSE, carefulfit = FALSE, finishsamples = 200)
)
```

```{r optimize-julia}
julia_map <- ctFit(
  datalong = dat, model = fit_model, backend = "julia",
  cores = 1, priors = TRUE, optimize = TRUE,
  optimcontrol = list(finishsamples = 200), verbose = 0
)
```

```{r optimize-compare}
rbind(stan = drift_interval(stan_map), julia = drift_interval(julia_map))
```

Each optimized fit above draws its `rawposterior` from a Hessian-based normal
approximation by default (`uncertainty = "hessian"`) -- the closest
optimized-fit analogue to the sampling intervals above, but still an
approximation around a single point rather than a description of the whole
density. The next section is about what else can go in that approximation's
place.

## The bits in between

An optimized fit does not have to settle for the default Hessian
approximation. `uncertainty` in `optimcontrol` (at fit time) or in
`ctOptimUncertainty()` (after the fact, reusing the existing optimum) selects
among several:

- `hessian`: finite-difference (stan) or exact forward-mode (julia) curvature
  at the optimum, with normal draws. The default, and the cheapest.
- `surrogate`: a local quadratic fit using gradient evaluations at points
  spread around the optimum (`max(4 * npars, 50)` by default). More stable
  than a single finite-difference Hessian when individual directions are
  noisy, still local.
- `opg`: an information estimate from the variability of per-subject score
  contributions, with no Hessian involved. Included as a diagnostic; it tends
  to disagree with the others most exactly where the Hessian's curvature
  information would have mattered.
- `sandwich`: Hessian bread with score-covariance meat -- a robust covariance
  more than a curvature approximation, useful as a misspecification check.
- `bootstrap`: a one-step score bootstrap using the Hessian as bread. Much
  cheaper than refitting, but score-based methods need enough independent
  subject-level score rows to be stable; with few subjects, expect a warning
  and a wider, noisier interval.
- `fullbootstrap`: resample subjects and fully re-optimize for each resample.
  The most literal bootstrap and the most expensive -- for `backend = 'julia'`
  it is the one method not available, because it needs the model rebuilt per
  resample rather than just re-evaluated at new data.
- `is`: importance sampling using the Hessian as the proposal covariance,
  with uncertainty read from the resulting weighted sample.

Every method except `fullbootstrap` works on both backends, reusing the same
underlying `ctOptimComputeUncertainty()` machinery either way -- the julia
path supplies its own exact Hessian and per-subject scores to the same code
rather than duplicating the numerics.

```{r uncertainty-gallery}
methods <- c("hessian", "surrogate", "opg", "sandwich", "bootstrap", "is")

gallery <- function(base_fit, backend_label) {
  do.call(rbind, lapply(methods, function(m) {
    updated <- ctOptimUncertainty(base_fit, uncertainty = m, finishsamples = 200, cores = 1)
    data.frame(backend = backend_label, method = m, t(drift_interval(updated)))
  }))
}

uncertainty_table <- rbind(gallery(stan_map, "stan"), gallery(julia_map, "julia"))
uncertainty_table
```

```{r uncertainty-plot, fig.height=4}
ggplot(uncertainty_table, aes(x = method, y = mean, ymin = X2.5., ymax = X97.5.,
                               colour = backend)) +
  geom_pointrange(position = position_dodge(width = 0.4)) +
  theme_bw() +
  labs(x = NULL, y = "drift, 95% interval")
```

With eight subjects the score-based methods (`opg`, `sandwich`, `bootstrap`)
warn about too few independent score rows and show it: wider and less stable
intervals than `hessian` or `surrogate`. That warning is informative rather
than a bug in the demonstration -- it is the method telling you its assumption
does not hold here. `fullbootstrap` is shown separately because it is
stan-only and slower even at a small resample count:

```{r fullbootstrap}
stan_fullboot <- ctOptimUncertainty(
  stan_map, uncertainty = "fullbootstrap", finishsamples = 8, cores = 1,
  control = list(bootstrapTol = 1e-3)
)
drift_interval(stan_fullboot)
```

### The state-explicit path has one route, not seven

Everything above assumes `intoverstates = TRUE`, the default: a Kalman filter
integrates the latent states out, and the fit's objective is the marginal
likelihood (or posterior) over the parameters alone. Setting
`intoverstates = FALSE` instead makes the states part of what is estimated --
useful with non-Gaussian indicators, where `backend = 'julia'` can integrate
the observation exactly rather than linearising it the way the filter does.

Optimizing that model finds the *joint* mode of parameters and states
together, which is not the same thing as the marginal maximum likelihood --
variance parameters are biased downward, in the same way a random-effects
model's variance is biased if the effects are treated as parameters to
optimize rather than integrated out. Every uncertainty method above except
`hessian` scores the marginal likelihood, which this fit did not maximise, so
`ctOptimUncertainty()` refuses them for an `intoverstates = FALSE` fit.
`hessian` still works because it profiles the enormous state block out of the
curvature (`ctsem_joint_hessian(..., profile = TRUE)` in the julia engine) --
but the profile can be close to flat, and on a model this small it is:

```{r state-explicit-optimize}
count_gen <- ctModel(
  type = "ct", n.latent = 1, n.manifest = 1,
  manifestNames = "c1", latentNames = "eta1", manifesttype = 3L,
  LAMBDA = matrix(1), DRIFT = matrix(-0.5), DIFFUSION = matrix(0.7),
  MANIFESTVAR = matrix(0), T0VAR = matrix(1), T0MEANS = matrix(0),
  CINT = matrix(0), MANIFESTMEANS = matrix(1), Tpoints = 6
)
count_dat <- data.frame(ctGenerate(count_gen, n.subjects = 8, Tpoints = 6, backend = "julia"))

count_fit_model <- ctModel(
  type = "ct", n.latent = 1, n.manifest = 1,
  manifestNames = "c1", latentNames = "eta1", manifesttype = 3L,
  LAMBDA = matrix(1), MANIFESTVAR = matrix(0), T0MEANS = matrix(0), CINT = matrix(0),
  MANIFESTMEANS = matrix("cmean"), Tpoints = 6
)
count_fit_model$pars$indvarying <- FALSE

state_explicit_opt <- ctFit(
  count_dat, count_fit_model, backend = "julia",
  intoverstates = FALSE, optimize = TRUE, priors = TRUE, verbose = 0
)
is.null(state_explicit_opt$estimate$se)
```

`manifesttype = 3` is a Poisson-linked count indicator: eight subjects and six
time points give the filter almost nothing to profile the state block down
to, so the profiled Hessian returns no usable standard errors at all rather
than merely wide ones. The documented way out is the same as the general
sampling-versus-optimizing choice above, applied to the states as well as the
parameters: fit with `optimize = FALSE` and let HMC sample the joint posterior
of parameters and states together, with no Gaussian assumption about the
states anywhere.

```{r state-explicit-sample}
state_explicit_samp <- ctFit(
  count_dat, count_fit_model, backend = "julia",
  intoverstates = FALSE, optimize = FALSE, priors = TRUE,
  chains = 1, iter = 120, verbose = 0
)
state_explicit_samp$estimate$se
dim(state_explicit_samp$estimate$states)
```

The sampled fit reports finite standard errors from the draws directly, and
`estimate$states` carries the sampled trajectory alongside the parameters --
the whole reason to make the states explicit rather than let the filter
integrate them away.

## With and without priors

`priors = TRUE` adds ctsem's raw-scale priors to the objective; `priors =
FALSE` (the `ctFit()` default) leaves them out. Combined with `optimize`, that
is four corners, and they mean four different things:

|              | `priors = TRUE`                          | `priors = FALSE`                    |
|--------------|-------------------------------------------|--------------------------------------|
| `optimize = TRUE`  | MAP: posterior mode, curvature around it   | MLE: likelihood maximum, curvature around it |
| `optimize = FALSE` | full posterior draws                       | likelihood-only draws -- see below   |

The first three corners are unremarkable and already shown above (`stan_map`
/ `julia_map` are MAP fits; an MLE fit is the same call with `priors = FALSE`;
`stan_sample` / `julia_sample` are posterior draws). The fourth is the one
users reach for without meaning to -- `priors = FALSE, optimize = FALSE` --
and both backends warn about it rather than refusing it outright:

```{r likelihood-only-sample}
stan_lik_sample <- ctFit(
  datalong = dat, model = fit_model, backend = "stan",
  cores = 1, priors = FALSE, optimize = FALSE,
  chains = 1, iter = 200, verbose = 0
)
range(stan_lik_sample$stanfit$rawposterior)
```

```{r likelihood-only-sample-julia}
julia_lik_sample <- ctFit(
  datalong = dat, model = fit_model, backend = "julia",
  cores = 1, priors = FALSE, optimize = FALSE,
  chains = 1, iter = 150, verbose = 0
)
range(julia_lik_sample$estimate$rawposterior)
```

Compare the raw-scale range here to the posterior sample earlier, which
stayed within a couple of units of zero on the same raw scale. With no prior
at all, the raw parameters have nothing constraining them outside what the
likelihood itself identifies, and for a model and sample size this small that
is not much. Run at these settings, the julia backend often names the
mechanism directly, with a warning that some sampled parameter "reached the
region where their transform is flat to machine precision... with no prior
the posterior is improper there and the chains wander rather than mix" --
suppressed from this rendering along with every other warning, but worth
reading in full the first time it appears in an interactive session. The stan
sampler shows the same problem less legibly: warm-up tends to take
substantially longer with `priors = FALSE` than with `priors = TRUE` on this
model, because the sampler is fighting geometry that a prior would have
tamed, and the resulting draws are still not trustworthy afterward.

None of this means `priors = FALSE` sampling is broken -- it is sampling
exactly what was asked for, the raw likelihood, and for a well identified
model with enough data it is a legitimate route to a frequentist-flavoured
posterior. It means that removing the prior removes the thing keeping the raw
scale bounded, and a small or weakly identified model will show you that
immediately rather than converging to something that merely looks plausible.
`priors = TRUE` is the safer default for sampling for exactly this reason,
whatever `ctFit()`'s own default is for optimizing.

## Summary

- Sampling and optimizing target the same density for a given `priors`
  setting; they differ in whether the answer is a set of draws or a point plus
  a local approximation.
- Around an optimum, `uncertainty` chooses which approximation: `hessian` is
  the default and cheapest, `surrogate` is a more stable local alternative,
  `opg`/`sandwich`/`bootstrap` use per-subject scores and need enough subjects
  to be stable, `fullbootstrap` (stan only) is the most literal and most
  expensive, and `is` turns the same curvature into an importance-sampling
  proposal.
- `intoverstates = FALSE` narrows this to one route at optimize time
  (`hessian`, profiling the state block out of the curvature) because
  everything else would score an objective the fit did not maximise; sampling
  instead of optimizing removes that restriction by sampling the states along
  with the parameters.
- Priors change what the objective *is*, not just how confident the reported
  uncertainty looks: MAP and MLE for optimizing, a real posterior or a
  likelihood-only draw for sampling. The likelihood-only corner is the one to
  be careful with -- it is supported, but a small or weakly identified model
  will show you exactly why priors are the default recommendation for
  sampling, by wandering.
