---
title: "Optimized ctsem Uncertainty Approximations"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Optimized ctsem Uncertainty Approximations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(ctsem)
library(ggplot2)
set.seed(12)
```

This example compares uncertainty approximations for an optimized ctsem fit.
All methods start from the same optimized maximum likelihood or maximum a
posteriori estimate and then refresh the approximate raw-parameter draws with
`ctOptimUncertainty()`.

For optimized fits, ctsem works on the raw unconstrained parameter scale.
Let $ \hat\theta $ be the optimized raw parameter vector and let
$\ell(\theta)$ be the active Stan log objective. For maximum likelihood fits
this is the log likelihood plus transform adjustments; for MAP fits it also
includes active priors. The same objective is used for log probabilities,
gradients, Hessians, and score contributions. The default optimized-fit
uncertainty is a local approximation around $ \hat\theta $. ctsem draws raw
parameter samples from that approximation and then maps them through the
transforms to obtain summaries on the usual ctsem parameter scales.

These draws are not full MCMC draws. They are useful when an optimized ML or
MAP fit is preferred for speed or stability, but their quality depends on the
local shape of the optimized objective and on the approximation selected.

## Methods

The uncertainty method is selected by `uncertainty` in `optimcontrol` during
`ctFit()`, or by `uncertainty` in `ctOptimUncertainty()` for an existing
optimized fit.

- `hessian`: finite-difference Hessian covariance with normal draws.
- `opg`: OPG-style score information approximation.
- `sandwich`: Hessian bread with score covariance meat.
- `bootstrap`: one-step score bootstrap draws.
- `fullbootstrap`: subject-resampling bootstrap with full re-optimization.
- `surrogate`: local quadratic surrogate around the optimum.

### Hessian

`uncertainty = "hessian"` estimates the local curvature of the active log
objective by finite differences of the gradient at the optimum. If $H$ is the
Hessian of $\ell(\theta)$, the approximate covariance is

$$
\Sigma_\theta \approx (-H)^{-1}.
$$

This is the closest optimized-fit analogue to the usual normal approximation
around an optimum. It is usually the right first choice when the optimum is
well identified and the objective is close to quadratic on the raw parameter
scale. It can be fragile when the Hessian is numerically unstable, parameters
are weakly identified, or the objective has strong skew, boundaries, or
multiple modes.

Relevant controls:

- `uncertaintyControl = list(hessianStep = ...)` changes the finite-difference
  step size.
- `uncertaintyControl = list(ridge = ...)` controls the minimum eigenvalue used
  when stabilising covariance matrices.
- `uncertaintyDraws = "normal"` or `draws = "normal"` draws from the resulting
  multivariate normal approximation.

### Surrogate

`uncertainty = "surrogate"` fits a local quadratic approximation using gradient
evaluations at points around the optimum. It still targets local curvature, but
uses multiple nearby evaluations instead of only one finite-difference Hessian.
This can be more stable than a direct Hessian when individual finite-difference
directions are noisy.

The design is dimension adjusted. By default, ctsem uses at least
`max(4 * npars, 50)` local points, where `npars` is the number of optimized raw
parameters. The initial point spread is also adjusted by dimension so that the
expected log-objective drop from the optimum remains local rather than growing
linearly with the number of parameters. Candidate points that are much too
close to the optimum or too far into the tails are dropped, and additional
points are sampled after global and parameterwise scale adjustments when
needed.

Surrogate uncertainty is still local. It will not discover remote modes and it
will not repair a poor optimum. It is most useful when the Hessian is noisy but
the objective is plausibly quadratic in the neighbourhood that matters.

Relevant controls:

- `uncertaintyControl = list(surrogateNpoints = ...)` sets the number of local
  design points. If omitted, the default scales with the number of raw
  parameters. Larger values are slower and usually more stable.
- `uncertaintyControl = list(surrogateScale = ...)` controls how far the design
  points are spread around the optimum, relative to the internally
  dimension-adjusted starting scale.
- `uncertaintyControl = list(initialCov = ...)` can provide a starting
  covariance. `ctOptimUncertainty()` uses the existing fit covariance by
  default when available.

### One-step bootstrap

`uncertainty = "bootstrap"` computes subject or case score contributions at the
optimum and uses a one-step update,

$$
\theta^* \approx \hat\theta + \Sigma_\theta s^*,
$$

where $\Sigma_\theta$ is the Hessian covariance and $s^*$ is a resampled
sum of centred score contributions. This preserves the local Hessian "bread"
and perturbs the estimate using score variability. The score contributions are
computed as Stan gradients of the same active objective used for fitting,
including transform adjustments and active priors when present.

This is the recommended bootstrap-style option for optimized fits. It is much
cheaper than refitting a model for every bootstrap sample, but it is still a
linear approximation. It can be useful when subject-level influence varies more
than the local Hessian approximation suggests. With few subjects, score-based
methods can be unstable because there are too few independent score
contributions to resample.

By default, `uncertaintyDraws = "auto"` and `draws = "auto"` use empirical
one-step bootstrap draws for `uncertainty = "bootstrap"`.

### Full Bootstrap

`uncertainty = "fullbootstrap"` resamples subjects with replacement and
re-optimizes the model for each bootstrap dataset. 

This is the most literal bootstrap option and is usually much slower than the
one-step bootstrap. It can capture some nonlinearity that the one-step update
misses, but it still depends on the optimizer finding the relevant local
optimum for each resampled dataset. It is mainly useful as a benchmark for the
faster approximations or when the linear one-step update is suspect.

For full bootstrap, `cores` is treated as the total core budget. By default,
ctsem runs one core per bootstrap refit and parallelizes over bootstrap
samples. Use `control = list(bootstrapFitCores = k)` when each refit should
split log-probability and gradient evaluations across subjects; the number of
parallel bootstrap workers is then reduced to avoid oversubscribing the
machine. `control = list(bootstrapTol = ...)` controls the mize L-BFGS
tolerance for each refit.

### Sandwich

`uncertainty = "sandwich"` also combines Hessian bread with score covariance
meat:

$$
\Sigma_\theta \approx (-H)^{-1} \left(\sum_i s_i s_i^\top\right) (-H)^{-1}.
$$

This is a robust covariance approximation rather than a direct local-curvature
approximation. It is mainly useful as a sensitivity check for misspecification
or heterogeneous subject contributions. It can be wider than the Hessian
curvature approximation, and that may be the intended frequentist behaviour
rather than a bug.

### OPG

`uncertainty = "opg"` uses score variability itself as an information estimate.
It is included as a diagnostic. The score contributions are gradients of the
same active objective used for fitting, so active prior contributions are
included. OPG differs because it does not use the local Hessian bread.
Consequently it can behave differently from Hessian, bootstrap, and sandwich
approximations when local curvature, including prior curvature in MAP fits, is
important. Prefer `bootstrap` or `sandwich` when score contributions are of
interest.

## Draws And Arguments

There are two separate decisions:

- `uncertainty` chooses the covariance or empirical approximation.
- `uncertaintyDraws` in `ctFit()` or `draws` in `ctOptimUncertainty()` chooses
  how approximate raw-parameter samples are generated from that approximation.

The available draw rules are:

- `"auto"`: empirical draws for `uncertainty = "bootstrap"` and
  `uncertainty = "fullbootstrap"`, normal draws otherwise.
- `"normal"`: multivariate normal draws from the selected covariance.
- `"empirical"`: empirical draws when the method provides them, otherwise falls
  back to normal draws.
- `"imis"`: importance sampling using the selected covariance as the proposal
  scale.

`finishsamples` controls the number of approximate raw-parameter samples stored in
`fit$stanfit$rawposterior` and then transformed for summaries. Increase it when
interval quantiles are noisy. `cores` controls both fitting and the uncertainty
work inside a single `ctFit()` call; for coverage simulations, use
`ctModelCoverage_check(cores = ..., fit_cores = ...)` to avoid accidental
nested parallelism.

Score-based methods use subject-level score contributions whenever there are
at least two subjects. This is usually the appropriate target for
multi-subject ctsem models because observations within a subject are not
independent. Single-subject models fall back to case-level scores with a
warning. Score-based methods also warn when there are fewer than ten
independent subject-level score rows, or when there are no more score rows
than raw parameters, because the covariance estimate is then unstable or rank
limited. `fullbootstrap` resamples subjects and therefore requires at least
two subjects; with fewer than ten subjects it runs with a warning because the
bootstrap distribution can be unstable. Bootstrap-style methods also require
at least two returned samples or refits so that a covariance can be estimated.

The uncertainty methods use whatever objective the fit used. With
`priors = FALSE`, this is the maximum likelihood objective plus transform
adjustments. With `priors = TRUE`, priors are part of the objective and are
included in log probabilities, gradients, Hessians, and score contributions.
In `scorecalc()`, ctsem distributes active prior contributions across
subject-level or case-level score components using `priormod`, so summed score
contributions remain on the same objective scale. The distinction between
methods is not whether priors are included in scores; they are. The distinction
is whether the approximation uses local curvature, as in Hessian, surrogate,
bootstrap, and sandwich methods, or only score variability, as in OPG.

Useful fit-time arguments are:

- `optimcontrol = list(uncertainty = "hessian")`
- `optimcontrol = list(uncertaintyDraws = "auto")`
- `optimcontrol = list(uncertaintyControl = list(...))`
- `optimcontrol = list(finishsamples = 1000)`
- `optimcontrol = list(is = TRUE, isESS = ..., isitersize = ...)` for the
  existing importance-sampling path. `is = TRUE` is equivalent to requesting
  importance-sampling draws.

Useful post-hoc arguments are:

- `ctOptimUncertainty(fit, uncertainty = "surrogate")`
- `ctOptimUncertainty(fit, draws = "normal")`
- `ctOptimUncertainty(fit, finishsamples = 1000, cores = 4)`
- `ctOptimUncertainty(fit, control = list(surrogateNpoints = 100))`
- `ctOptimUncertainty(fit, uncertainty = "fullbootstrap", control = list(bootstrapFitCores = 1))`

Use post-hoc updates when the optimized point estimate is satisfactory but the
uncertainty approximation needs comparison or refinement. `ctOptimUncertainty()`
is for optimized `ctFit` objects. It is not needed for full MCMC fits,
where uncertainty is already represented by the sampled posterior.

## Generate Data

```{r}
n_subjects <- 10
tpoints <- 16

gen_model <- ctModel(
  type = "omx",
  Tpoints = tpoints,
  n.latent = 1,
  n.manifest = 1,
  LAMBDA = matrix(1),
  T0MEANS = matrix(0),
  T0VAR = matrix(0.6),
  DRIFT = matrix(-0.35),
  DIFFUSION = matrix(0.5),
  CINT = matrix(0.15),
  MANIFESTMEANS = matrix(0),
  MANIFESTVAR = matrix(0.25)
)

dat <- do.call(rbind, lapply(seq_len(n_subjects), function(id) {
  d <- data.frame(ctGenerate(
    ctmodelobj = gen_model,
    n.subjects = 1,
    burnin = 0,
    dtmean = 0.4,
    logdtsd = 0.1
  ))
  d$id <- id
  d
}))
dat <- dat[order(dat$id, dat$time), ]
```

The fitted model estimates a continuous-time intercept, drift, and diffusion
for a single latent process with fixed measurement scale.

```{r}
fit_model <- ctModel(
  type = "ct",
  Tpoints = tpoints,
  n.latent = 1,
  n.manifest = 1,
  LAMBDA = matrix(1),
  MANIFESTMEANS = matrix(0),
  CINT = matrix("cint"),
  DRIFT = matrix("drift"),
  DIFFUSION = matrix("diffusion")
)
```

## Fit Once

The initial fit uses the existing Hessian-based optimized-fit approximation.
Subsequent sections reuse its optimized estimate and compiled model.

```{r}
base_fit <- ctFit(
  datalong = dat,
  model = fit_model,
  cores = 1,
  priors = TRUE,
  optimize = TRUE,
  optimcontrol = list(
    stochastic = FALSE,
    carefulfit = FALSE,
    finishsamples = 120,
    uncertainty = "hessian"
  )
)
```

## Compare Methods

```{r}
raw_widths <- function(fit) {
  raw <- fit$stanfit$rawposterior
  q <- apply(raw, 2, quantile, probs = c(.025, .975), na.rm = TRUE)
  data.frame(
    parameter = paste0("raw", seq_len(ncol(raw))),
    width = as.numeric(q[2, ] - q[1, ])
  )
}

time_update <- function(method, draws = "auto", control = list(), finish = 120) {
  elapsed <- system.time({
    fit <- ctOptimUncertainty(
      base_fit,
      uncertainty = method,
      draws = draws,
      finishsamples = finish,
      cores = 1,
      control = control
    )
  })[["elapsed"]]
  list(method = method, seconds = elapsed, fit = fit)
}

results <- list(
  time_update("hessian"),
  time_update("opg"),
  time_update("sandwich"),
  time_update("bootstrap"),
  time_update("fullbootstrap", finish = 20, control = list(bootstrapTol = 1e-4)),
  time_update("surrogate", control = list(surrogateNpoints = 30, surrogateScale = .6))
)

timings <- data.frame(
  method = vapply(results, `[[`, character(1), "method"),
  seconds = vapply(results, `[[`, numeric(1), "seconds")
)

widths <- do.call(rbind, lapply(results, function(x) {
  data.frame(method = x$method, raw_widths(x$fit))
}))
```

```{r}
timings
```

```{r}
ggplot(timings, aes(x = method, y = seconds)) +
  geom_col(fill = "grey45") +
  theme_bw() +
  labs(x = NULL, y = "Elapsed seconds")
```

```{r}
ggplot(widths, aes(x = parameter, y = width, colour = method, group = method)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  labs(x = "Raw parameter", y = "95% interval width")
```

## Direct Use During Fitting

The same uncertainty types can be requested directly during `ctFit()` through
`optimcontrol`.

```{r, eval=FALSE}
fit_surrogate <- ctFit(
  datalong = dat,
  model = fit_model,
  cores = 1,
  priors = TRUE,
  optimize = TRUE,
  optimcontrol = list(
    stochastic = FALSE,
    carefulfit = FALSE,
    finishsamples = 120,
    uncertainty = "surrogate",
    uncertaintyControl = list(surrogateNpoints = 30)
  )
)
```

For fit-time use, select the method with `uncertainty`. The draw rule defaults
to `uncertaintyDraws = "auto"`, which uses empirical one-step draws for
`uncertainty = "bootstrap"` and normal draws otherwise.

## Coverage Simulation

`ctModelCoverage_check()` can run a small simulation study by first fitting the
model once, generating repeated datasets from that fitted model, and then
refitting each generated dataset under a set of named fitting options. This is
more expensive than the single-dataset comparison above, so the chunk below is
not evaluated during ordinary rendering. Set `eval=TRUE` and increase `niter`
for a real coverage run.

```{r, eval=FALSE}
coverage_fit_args <- list(
  hessian = list(
    priors = TRUE,
    optimize = TRUE,
    optimcontrol = list(
      stochastic = FALSE,
      carefulfit = FALSE,
      finishsamples = 100,
      uncertainty = "hessian"
    )
  ),
  opg = list(
    priors = TRUE,
    optimize = TRUE,
    optimcontrol = list(
      stochastic = FALSE,
      carefulfit = FALSE,
      finishsamples = 100,
      uncertainty = "opg"
    )
  ),
  sandwich = list(
    priors = TRUE,
    optimize = TRUE,
    optimcontrol = list(
      stochastic = FALSE,
      carefulfit = FALSE,
      finishsamples = 100,
      uncertainty = "sandwich"
    )
  ),
  bootstrap = list(
    priors = TRUE,
    optimize = TRUE,
    optimcontrol = list(
      stochastic = FALSE,
      carefulfit = FALSE,
      finishsamples = 100,
      uncertainty = "bootstrap"
    )
  ),
  surrogate = list(
    priors = TRUE,
    optimize = TRUE,
    optimcontrol = list(
      stochastic = FALSE,
      carefulfit = FALSE,
      finishsamples = 100,
      uncertainty = "surrogate",
      uncertaintyControl = list(
        surrogateNpoints = 30,
        surrogateScale = .6
      )
    )
  )
)

coverage_check <- ctModelCoverage_check(
  initialData = dat,
  fitting_model = fit_model,
  niter = 1000,
  fit_args = coverage_fit_args,
  cores = 20,
  fit_cores = 1,
  plot_every = 20
)

coverage_summary <- unique(data.frame(
  coverage_check$results[, c("type", "par", "coverage"), with = FALSE]
))
coverage_summary
```

The returned object also contains the final plots:

```{r, eval=FALSE}
coverage_check$bias_plot
coverage_check$coverage_plot
```

For publication-scale checks, use substantially larger `niter`, keep
`finishsamples` large enough for stable interval estimates, and use separate
sessions or cores for the outer simulation rather than increasing `ctFit()`
cores inside each replicate.
