---
title: "Treating a rating scale as continuous"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Treating a rating scale as continuous}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
  fig.height = 4, message = FALSE, warning = FALSE)
```

```{r libs}
library(ctsem)
library(ggplot2)
```

A rating scale can be modelled two ways in ctsem: `manifesttype = 0` puts a
Gaussian residual on the observed code, `manifesttype = 2` puts a cumulative
logit on it. Ordinal indicators require `backend = 'julia'`; the other
indicator types are covered in the measurement models vignette.

To see what the first choice changes, we simulate from a known process,
categorise it, and fit the same dynamic model both ways. Because the data are
simulated we can also fit the uncategorised latent states, which gives a
ceiling: whatever that fit fails to recover is due to the design and the
sample, not to categorisation.

Code that only generates or draws is folded throughout; unfold it if you want
to reproduce the figures.

# The data

Two Ornstein-Uhlenbeck processes, one five point item each. `eta1` affects
`eta2` and not the reverse, the innovations correlate at 0.4, and four
thresholds at $\pm1$ and $\pm2.8$ cut each process into categories.

```{r data}
#| code-fold: true
DRIFT <- matrix(c(-0.5, 0,
                  0.3, -0.5), 2, 2, byrow = TRUE)   # eta1 -> eta2 is 0.3
diffsd <- 2.2
diffcor <- 0.4
thresholds <- c(-2.8, -1.0, 1.0, 2.8)

# covmattransform='z' makes a covariance matrix's off-diagonal exactly
# atanh(correlation), so the generating values can be written down.
zmat <- function(sd1, sd2, r) matrix(c(sd1, 0, atanh(r), sd2), 2, 2, byrow = TRUE)

# Stationary covariance, solving A S + S A' + Q = 0, used for T0VAR so that
# every occasion is a draw from the same distribution.
Q <- diag(diffsd, 2) %*% matrix(c(1, diffcor, diffcor, 1), 2) %*% diag(diffsd, 2)
S <- matrix(solve(kronecker(diag(2), DRIFT) + kronecker(DRIFT, diag(2)),
  -as.vector(Q)), 2, 2)
latentsd <- sqrt(diag(S))

gen <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
  manifestNames = c('eta1', 'eta2'), latentNames = c('eta1', 'eta2'),
  LAMBDA = diag(1, 2), DRIFT = DRIFT,
  DIFFUSION = zmat(diffsd, diffsd, diffcor),
  T0VAR = zmat(latentsd[1], latentsd[2], S[1, 2] / prod(latentsd)),
  T0MEANS = matrix(0, 2, 1), CINT = matrix(0, 2, 1),
  MANIFESTMEANS = matrix(0, 2, 1),
  MANIFESTVAR = diag(1e-6, 2),          # no measurement error: manifest = state
  Tpoints = 30)
gen$covmattransform <- 'z'

# backend='julia' because the R generator reads DIFFUSION and T0VAR as square
# roots whatever covmattransform says, while the julia generator applies the
# declared transform, as ctFit() does.
set.seed(2)
d <- data.frame(ctGenerate(gen, n.subjects = 150, Tpoints = 30, dtmean = 0.7,
  backend = 'julia'))

# The cumulative logit: the response falls in category k or below when the
# state plus a logistic disturbance is below threshold k.
ordinalise <- function(eta)
  1 + rowSums(outer(eta + rlogis(length(eta)), thresholds, '>'))
set.seed(502)
d$y1 <- ordinalise(d$eta1)
d$y2 <- ordinalise(d$eta2)
```

One subject. The line is the latent process, the dotted lines are the
thresholds, and the points are the recorded answers at the middle of the band
each category stands for:

```{r overlay, fig.height = 4}
#| code-fold: true
mid <- c(-3.8, -1.9, 0, 1.9, 3.8)             # band midpoints, for display
s <- d[d$id == 1, ]

ggplot(s, aes(time)) +
  geom_hline(yintercept = thresholds, linetype = 3, colour = 'grey55') +
  geom_line(aes(y = eta1, colour = 'latent eta1')) +
  geom_point(aes(y = mid[y1], colour = 'recorded category y1'), size = 1.9) +
  scale_y_continuous(sec.axis = sec_axis(~., breaks = mid, labels = 1:5,
    name = 'category')) +
  scale_colour_manual(values = c('latent eta1' = 'grey35',
    'recorded category y1' = '#c0392b')) +
  theme_bw() + theme(legend.position = 'bottom') +
  labs(x = 'time', y = 'latent scale', colour = NULL)
```

The points follow the line, but only to the resolution of the bands, and they
cross a threshold early or late depending on the disturbance. Both items use
the whole scale, with no category so rare that it carries nothing:

```{r categoryuse, fig.height = 2.4}
#| code-fold: true
freq <- rbind(data.frame(item = 'y1', category = factor(d$y1)),
  data.frame(item = 'y2', category = factor(d$y2)))

ggplot(freq, aes(category)) + geom_bar() + facet_wrap(~item) +
  theme_bw() + labs(x = NULL, y = NULL)
```

# Fitting it as ordinal

This is the whole of it. Two arguments do the work: `manifesttype = 2` asks
for the cumulative logit, and `ncategories` says how many thresholds to
estimate, which the model has to know before it sees the data.

```{r ordinalfit}
ordinalmodel <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
  manifestNames = c('y1', 'y2'), latentNames = c('eta1', 'eta2'),
  manifesttype = c(2, 2),             # ordinal
  ncategories = c(5, 5),              # five point scale
  LAMBDA = diag(1, 2),
  CINT = matrix(0, 2, 1), T0MEANS = matrix(0, 2, 1))

ordinalfit <- ctFit(d[, c('id', 'time', 'y1', 'y2')], ordinalmodel,
  backend = 'julia', cores = 1)
```

Nothing is said about `MANIFESTVAR`. The cumulative logit already supplies the
randomness, so a residual variance on top of it is not identified, and
`ctFit()` fixes the whole matrix itself and says so -- including the
off-diagonal, which the default leaves free as a correlated measurement error
between the two items. Setting it to zero by hand changes nothing.

`summary()` reports the thresholds on the latent scale, one row per category
boundary:

```{r thresholds}
summary(ordinalfit)$thresholds
```

Those are recovered against generating values of −2.8, −1, 1 and 2.8.

# Fitting it as continuous, and the ceiling

The Gaussian version is the same call with `manifesttype = 0`, no
`ncategories`, and a free residual variance per item instead of thresholds.
The ceiling fit is that same Gaussian model applied to the uncategorised
states, which is only possible because the data are simulated.

```{r otherfits}
#| code-fold: true
gaussianmodel <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
  manifestNames = c('y1', 'y2'), latentNames = c('eta1', 'eta2'),
  LAMBDA = diag(1, 2), CINT = matrix(0, 2, 1), T0MEANS = matrix(0, 2, 1),
  MANIFESTVAR = matrix(c('mvar_y1', 0, 0, 'mvar_y2'), 2, 2))

# The ceiling data are the latent states themselves, so there is no
# measurement error to estimate: leaving it free puts a parameter on the
# boundary at zero and the fit reports a direction it cannot identify.
ceilingmodel <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
  manifestNames = c('c1', 'c2'), latentNames = c('eta1', 'eta2'),
  LAMBDA = diag(1, 2), CINT = matrix(0, 2, 1), T0MEANS = matrix(0, 2, 1),
  MANIFESTVAR = diag(0, 2))
ceilingdata <- setNames(d[, c('id', 'time', 'eta1', 'eta2')],
  c('id', 'time', 'c1', 'c2'))

gaussianfit <- ctFit(d[, c('id', 'time', 'y1', 'y2')], gaussianmodel,
  backend = 'julia', cores = 1)
ceilingfit <- ctFit(ceilingdata, ceilingmodel, backend = 'julia', cores = 1)
```

# What should and should not agree

The two fits place the latent processes on different scales. Both are set by
the same convention -- `LAMBDA` is fixed at 1, so the latent takes its units
from the indicator -- but the units differ: the Gaussian fit measures `eta1`
in answer codes, the ordinal fit in logits, because the cumulative logit's own
scale is fixed and the latent is measured against it.

The thresholds do not fix that scale; they would rescale along with the
process. What pins it is the link. Free the loadings as well and nothing pins
it at all, which is the identification problem the item response vignette
deals with.

A parameter can therefore only be compared across the two fits if it is
invariant to rescaling the latent processes.

**Expected to agree**

* *Auto effects* (`drift_eta1`, `drift_eta2`). Their unit is inverse time, and
  a rate of reversion does not change when the process is measured in other
  units.
* *The diffusion correlation*, being a correlation.
* *The standardised cross effect*, the effect in stationary standard
  deviations, $\tilde{a}_{21} = a_{21}\,\text{sd}(\eta_1)/\text{sd}(\eta_2)$.

**Not expected to agree**

* *The raw cross effect* `drift_eta2_eta1`, whose unit is `eta2` per unit of
  `eta1` per unit time. Note this is already true between the two processes
  within a single fit: `eta2` is the more variable because it accumulates
  `eta1`, so the same effect takes different numbers on differently scaled
  processes.
* *`DIFFUSION` and `T0VAR` on their own*, and the residual variances.
* *`MANIFESTMEANS`*, a location on the latent scale in the Gaussian fit and
  the thresholds' anchor in the ordinal one.

Only the ordinal fit's scale means anything outside the questionnaire, so
where the raw quantities differ it is the ordinal fit to report.

# Comparison

`ctDiscretePars()` turns each fit's `DRIFT` into the regressions it implies at
every time interval. Standardised, those are comparable across the three:

```{r crosscurve}
#| code-fold: true
times <- seq(0, 6, .1)

crosscurve <- function(fit, label) {
  a <- ctDiscretePars(fit, times = times, standardise = TRUE, cores = 1)
  data.frame(time = times, model = label,
    value = apply(a, c(3, 4, 5), median, na.rm = TRUE)[, 2, 1])
}

curves <- rbind(
  crosscurve(ceilingfit, 'uncategorised (ceiling)'),
  crosscurve(gaussianfit, 'gaussian on the codes'),
  crosscurve(ordinalfit, 'ordinal'),
  data.frame(time = times, model = 'truth', value = sapply(times,
    function(u) expm::expm(DRIFT * u)[2, 1] * latentsd[1] / latentsd[2])))

ggplot(curves, aes(time, value, colour = model, linetype = model)) +
  geom_line(linewidth = .9) + theme_bw() +
  theme(legend.position = 'bottom') +
  labs(x = 'time interval', y = 'standardised cross effect, eta1 -> eta2',
    colour = NULL, linetype = NULL)
```

The auto effect and the diffusion correlation come from
`summary(fit)$parmatrices`, which carries `DIFFUSIONcov`, the innovation
covariance, and `asymDIFFUSIONcov`, the stationary covariance the model
implies:

```{r table}
#| code-fold: true
comparable <- function(fit, label) {
  pm <- summary(fit)$parmatrices
  cell <- function(mat, i, j) pm$Mean[pm$matrix == mat & pm$row == i & pm$col == j]
  data.frame(model = label,
    drift_eta1 = summary(fit)$popmeans['drift_eta1', 'mean'],
    latentsd1 = sqrt(cell('asymDIFFUSIONcov', 1, 1)),
    diffcor = cell('DIFFUSIONcov', 1, 2) /
      sqrt(cell('DIFFUSIONcov', 1, 1) * cell('DIFFUSIONcov', 2, 2)))
}

out <- rbind(
  comparable(ceilingfit, 'uncategorised (ceiling)'),
  comparable(gaussianfit, 'gaussian on the codes'),
  comparable(ordinalfit, 'ordinal'),
  data.frame(model = 'truth', drift_eta1 = DRIFT[1, 1],
    latentsd1 = latentsd[1], diffcor = diffcor))
cbind(out[1], round(out[-1], 3))
```

`latentsd1` is in answer-code units for the Gaussian fit and in logits for the
ordinal fit, so the two fitted columns are not comparable with each other. The
ordinal one *is* comparable with the truth, because the link fixes its units:
it estimates a real quantity rather than a convention, and checking it is
worth doing before reading anything else.

Categorisation costs little here. That is expected under the assumptions used:
categorisation is a monotone transformation plus noise that is independent
across items and occasions given the latent state, and `MANIFESTVAR` in the
Gaussian model absorbs that noise. The Gaussian model still gets the latent
scale wrong, which is why the raw cross effect cannot be read across the two
fits, but the scale-invariant quantities are recovered.
