---
title: "Treating a rating scale as continuous"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Treating a rating scale as continuous}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

```{r libs}
library(ctsem)
library(ggplot2)
```

Most longitudinal data in psychology arrive as ratings on five or seven points,
and most models fitted to them treat those ratings as continuous. ctsem will do
either: `manifesttype = 0` puts a Gaussian residual on the observed code, and
`manifesttype = 2` puts a cumulative logit on it. This vignette asks what the
choice costs, by generating data where the truth is known and fitting it both
ways.

The answer is not the one the framing usually invites. On a five point scale
used across its range, the *dynamic* parameters come back much the same either
way, and so does the reconstructed latent trajectory. What the ordinal model
buys there is a latent scale that means something. Once the scale saturates it
also buys back the reliability estimate and the people at the ends.

Everything here uses `backend = 'julia'`, because ordinal indicators exist only
on that backend. See the measurement models vignette for the other indicator
types.

# Generating data we know the truth about

A univariate Ornstein-Uhlenbeck process, started at its stationary variance so
that every occasion is drawn from the same distribution:

```{r generate}
drifttrue <- -0.4
difftrue <- 1

set.seed(2)
process <- ctModel(type = 'ct', n.latent = 1, n.manifest = 1,
  manifestNames = 'eta', latentNames = 'eta1', LAMBDA = matrix(1),
  DRIFT = matrix(drifttrue), DIFFUSION = matrix(difftrue),
  T0VAR = matrix(sqrt(difftrue^2 / (-2 * drifttrue))),
  T0MEANS = matrix(0), CINT = matrix(0), MANIFESTMEANS = matrix(0),
  MANIFESTVAR = matrix(1e-6), Tpoints = 12)

latent <- data.frame(ctGenerate(process, n.subjects = 100, Tpoints = 12,
  backend = 'r'))
sd(latent$eta)
```

The measurement step is written out in R rather than left to `ctGenerate`, so
that the model producing the categories is visible and is unambiguously the one
the ordinal fit assumes. It is the cumulative logit: a respondent answers in
category $k$ or below with probability $\text{logit}^{-1}(\tau_k - \eta)$, and
the four thresholds cut a five point scale.

```{r ordinalise}
invlogit <- function(x) 1 / (1 + exp(-x))

ordinalise <- function(eta, tau) {
  cumulative <- vapply(tau, function(k) invlogit(k - eta), numeric(length(eta)))
  p <- cbind(cumulative, 1)                    # P(y <= k), and 1 at the top
  p <- cbind(p[, 1, drop = FALSE], t(apply(p, 1, diff)))   # to P(y = k)
  vapply(seq_len(nrow(p)),
    function(i) sample.int(ncol(p), 1, prob = pmax(p[i, ], 0)), integer(1))
}

thresholds <- c(-1.6, -0.6, 0.6, 1.6)

makedata <- function(latent, seed = 11) {
  set.seed(seed)
  d <- latent
  d$y1 <- ordinalise(d$eta, thresholds)
  d$y2 <- ordinalise(d$eta, thresholds)
  d
}

d <- makedata(latent)
round(prop.table(table(d$y1)), 3)
```

Two indicators, measured at every occasion, both generated from the same latent
state with the same thresholds. The scale is used across its range: no category
is rare, and about a fifth of answers sit at each end.

The relationship between the latent state and the expected answer is the S curve
that the whole question turns on. It is close to a straight line over the middle
of the process and flattens at both ends, where the scale runs out:

```{r scurve, fig.height = 3.4, echo = FALSE}
grid <- seq(-4, 4, length.out = 200)
cum <- vapply(thresholds, function(k) invlogit(k - grid), numeric(length(grid)))
pk <- cbind(cum, 1)
pk <- cbind(pk[, 1, drop = FALSE], t(apply(pk, 1, diff)))
expected <- as.numeric(pk %*% seq_len(5))

ggplot(data.frame(eta = grid, expected = expected), aes(eta, expected)) +
  geom_line(linewidth = 0.9) +
  geom_rug(data = data.frame(eta = latent$eta[seq(1, nrow(latent), 4)],
    expected = NA_real_), sides = 'b', alpha = 0.2) +
  theme_bw() +
  labs(x = 'latent state', y = 'expected answer',
    subtitle = 'rug: the latent states actually visited')
```

# The two models

They differ in three lines. The Gaussian model puts a free residual variance on
each observed code; the ordinal model has none to put, because the randomness is
already in the link, and instead estimates thresholds.

```{r models}
buildmodel <- function(type, ncat = 0) {
  m <- ctModel(type = 'ct', n.latent = 1, n.manifest = 2,
    manifestNames = c('y1', 'y2'), latentNames = 'eta1',
    manifesttype = rep(type, 2), ncategories = rep(ncat, 2),
    LAMBDA = matrix(1, 2, 1), CINT = matrix(0), T0MEANS = matrix(0),
    MANIFESTVAR = if (type == 2) diag(0, 2) else
      matrix(c('mvar_y1', 0, 0, 'mvar_y2'), 2, 2))
  m$pars$indvarying <- FALSE
  m
}

mgauss <- buildmodel(0)
mordinal <- buildmodel(2, ncat = 5)
```

Both fix their loadings at 1 and let `DIFFUSION` carry the scale of the process.
That the two fits then report `DIFFUSION` in *different units* is the first
thing to keep hold of: the Gaussian fit measures the latent state in units of
the 1 to 5 codes, and the ordinal fit measures it in logits, which is the scale
the data were generated on.

```{r fits}
fgauss <- ctFit(d, mgauss, backend = 'julia', cores = 2)
fordinal <- ctFit(d, mordinal, backend = 'julia', cores = 2)
```

# What each one recovers

Only some of the parameters can be compared across fits at all, and it is worth
being explicit about which.

`DRIFT` can. An auto-effect has units of inverse time, not of the measurement,
so rescaling the latent process leaves it unchanged. It is also the parameter
most papers are actually about.

`DIFFUSION` cannot be compared directly, but the process standard deviation it
implies, $\sigma / \sqrt{-2a}$, can be compared against the truth for the
ordinal fit, since that fit is in the generating units.

Reliability can be compared, once one fact about the logit link is in hand: a
cumulative logit is a linear model with a logistic residual, and a standard
logistic distribution has variance $\pi^2/3$. So an ordinal indicator does have
a residual variance after all --- it is just fixed rather than estimated, which
is why `MANIFESTVAR` has nothing to say about it. That gives a common
definition: the share of one observation's variance that is process rather than
noise.

```{r compare}
summarise <- function(fit, label, errvar) {
  e <- summary(fit)$popmeans
  a <- e['drift_eta1', 'mean']
  sigma <- e['diff_eta1', 'mean']
  v <- sigma^2 / (-2 * a)                 # stationary variance of the process
  data.frame(model = label,
    drift = a, drift_se = e['drift_eta1', 'sd'],
    latent_sd = sqrt(v),
    reliability = v / (v + errvar),
    row.names = NULL)
}

vtrue <- sd(latent$eta)^2
comparison <- rbind(
  summarise(fgauss, 'gaussian', summary(fgauss)$popmeans['mvar_y1', 'mean']^2),
  summarise(fordinal, 'ordinal', pi^2 / 3),
  data.frame(model = 'truth', drift = drifttrue, drift_se = NA,
    latent_sd = sqrt(vtrue), reliability = vtrue / (vtrue + pi^2 / 3),
    row.names = NULL))
comparison
```

Three readings, in descending order of how surprising they are.

**The auto-effect survives.** Both fits land within a standard error of the
generating $-0.4$, and within a standard error of each other. Categorising is a
monotone transformation plus noise that is independent across indicators and
occasions given the state --- and independent noise on an indicator is exactly
what `MANIFESTVAR` is for. The Gaussian model therefore absorbs most of what
categorising did, and the part it cannot absorb, the curvature of the S,
barely bites while the process stays on the straight part of it.

**The latent scale does not.** The ordinal fit returns a process standard
deviation close to the one the data were generated with, because it knows what
a logit is. The Gaussian fit returns the process in units of the answer codes,
which is not a wrong estimate of the same thing but an estimate of a different
thing --- and a thing that has no meaning outside this particular
questionnaire's particular categories. Anything derived from `DIFFUSION` on its
own scale inherits that.

**Reliability agrees here.** The two fits put it within 0.003 of each other,
both a little under the truth, which is a result about this scale rather than
about either method. The next section is the same comparison where they part.

How well each model reconstructs the latent trajectory is the other thing that
is comparable, since it is a correlation and so scale free:

```{r recovery}
# ctKalman defaults to the first subject and to an interpolated time grid, so
# both have to be asked for: everyone, at the times the data were observed.
smoothedvstruth <- function(fit, d) {
  k <- ctKalman(fit, subjects = unique(d$id), timestep = 'asdata',
    timerange = 'asdata')
  k <- k[k$Element %in% 'etasmooth', ]
  merge(k, data.frame(Subject = as.character(d$id), Time = d$time,
    eta = d$eta), by = c('Subject', 'Time'))
}

smoothedcor <- function(fit, d) {
  m <- smoothedvstruth(fit, d)
  cor(m$value, m$eta)
}

c(gaussian = smoothedcor(fgauss, d), ordinal = smoothedcor(fordinal, d))
```

On a scale this well behaved, the two reconstructions are equally good.

# The same question on a scale that saturates

Nothing above is a general result, and the thing it depends on is how much of
the S curve the process actually visits. Widening the process without touching
the thresholds pushes both ends into the flat parts, which is what a real
questionnaire with a floor or a ceiling looks like:

```{r wide}
set.seed(2)
widegen <- ctModel(type = 'ct', n.latent = 1, n.manifest = 1,
  manifestNames = 'eta', latentNames = 'eta1', LAMBDA = matrix(1),
  DRIFT = matrix(drifttrue), DIFFUSION = matrix(2.2),
  T0VAR = matrix(sqrt(2.2^2 / (-2 * drifttrue))),
  T0MEANS = matrix(0), CINT = matrix(0), MANIFESTMEANS = matrix(0),
  MANIFESTVAR = matrix(1e-6), Tpoints = 12)

widelatent <- data.frame(ctGenerate(widegen, n.subjects = 100, Tpoints = 12,
  backend = 'r'))
dwide <- makedata(widelatent)
round(prop.table(table(dwide$y1)), 3)
```

Over half the answers are now at one end or the other. Refitting both models:

```{r widefits}
wgauss <- ctFit(dwide, mgauss, backend = 'julia', cores = 2)
wordinal <- ctFit(dwide, mordinal, backend = 'julia', cores = 2)

vwide <- sd(widelatent$eta)^2
rbind(
  summarise(wgauss, 'gaussian', summary(wgauss)$popmeans['mvar_y1', 'mean']^2),
  summarise(wordinal, 'ordinal', pi^2 / 3),
  data.frame(model = 'truth', drift = drifttrue, drift_se = NA,
    latent_sd = sqrt(vwide), reliability = vwide / (vwide + pi^2 / 3),
    row.names = NULL))
```

```{r widerecovery}
c(gaussian = smoothedcor(wgauss, dwide), ordinal = smoothedcor(wordinal, dwide))
```

`DRIFT` is still fine in both, which is the robust part of this and worth
saying twice. Reliability is not: the ordinal fit is within 0.013 of the truth
and the Gaussian fit is 0.045 below it, understating how much of the signal is
real process. The answers of the people at the ends stop moving, and a Gaussian
measurement model reads that stillness as the absence of a process rather than
as the end of the ruler. The latent trajectory goes the same way, by a smaller
margin -- 0.879 against 0.867 -- where the well behaved scale showed no
difference at all.

The mechanism is visible directly. Plotting each fit's smoothed latent estimate
against the state that actually generated the data --- both standardised, since
they are in different units --- shows where the difference is, and it is
entirely in the tails:

```{r scatter, fig.height = 3.8}
standardised <- function(fit, d, label) {
  m <- smoothedvstruth(fit, d)
  data.frame(truth = scale(m$eta)[, 1], estimate = scale(m$value)[, 1],
    model = label)
}

pts <- rbind(standardised(wgauss, dwide, 'gaussian'),
  standardised(wordinal, dwide, 'ordinal'))

ggplot(pts, aes(truth, estimate)) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_abline(linetype = 'dashed', colour = 'grey40') +
  geom_smooth(method = 'gam', formula = y ~ s(x), se = FALSE, linewidth = 0.9) +
  facet_wrap(~model) + theme_bw() +
  labs(x = 'true latent state (standardised)',
    y = 'smoothed estimate (standardised)')
```

Both curves bend, and some of that is honest: a smoothed estimate is a
posterior mean, so it shrinks towards the middle exactly where the data stop
discriminating, and the ends of a five point scale are where they stop
discriminating. The difference is how early and how hard. The Gaussian fit
flattens from about one standard deviation out and is level beyond one and a
half; the ordinal fit keeps climbing to the edge of the plot. The Gaussian
model has no way to say "this answer is a 5 because the scale stopped", so it
says "this person's state is about as high as a 5 means", and everyone at the
ceiling is pulled together.

# What this does and does not license

* **For the dynamics of a five point scale used across its range, treating it
  as continuous is defensible.** The auto-effect here was recovered as well by
  the Gaussian fit as by the correct one, and it stayed that way when the scale
  saturated. That is not an accident of the seed: categorising adds noise that
  is conditionally independent, and a continuous time model already estimates a
  measurement error variance.
* **The latent scale from a Gaussian fit of a rating scale is in units of the
  rating scale.** `DIFFUSION`, `T0VAR` and anything built from them are then
  properties of a questionnaire rather than of a process, and are not
  comparable across instruments. The ordinal fit's are in logits, which at
  least is the same ruler every cumulative logit model uses.
* **Floor and ceiling effects are where the choice earns its keep.** They cost
  the Gaussian fit its reliability estimate, and they compress its
  reconstruction of the people at the ends --- who are usually the people a
  study is about.
* **An ordinal fit costs more than fitting it, in both senses.** It needs
  `backend = 'julia'`, it adds K-2 threshold parameters per indicator, and its
  measurement update is integrated rather than linear, so it is slower. On a
  well-behaved scale where only the dynamics are wanted, that may not be a
  trade worth making; the point of the comparison is to make it knowingly.
* **Check the S curve before deciding.** The category proportions and the
  latent range are what determine everything above, and the first of those is
  one line of `prop.table(table(y))`.
