---
title: "Treating a rating scale as continuous"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Treating a rating scale as continuous}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

```{r libs}
library(ctsem)
library(ggplot2)
```

A rating scale can be modelled two ways in ctsem: `manifesttype = 0` puts a
Gaussian residual on the observed code, `manifesttype = 2` puts a cumulative
logit on it. Ordinal indicators require `backend = 'julia'`; the other
indicator types are covered in the measurement models vignette.

To see what the first choice changes, we simulate from a known process,
categorise it, and fit the same dynamic model both ways. Because the data are
simulated we can also fit the uncategorised latent states, which gives a
ceiling: whatever that fit fails to recover is due to the design and the
sample, not to categorisation.

# Data

Two Ornstein-Uhlenbeck processes with one item each. `eta1` affects `eta2` and
not the reverse, and the innovations correlate, so the pair has both a cross
effect and a contemporaneous association.

```{r data}
DRIFT <- matrix(c(-0.5, 0,
                  0.3, -0.5), 2, 2, byrow = TRUE)   # eta1 -> eta2 is 0.3
diffsd <- 2.2                                       # innovation sd, both
diffcor <- 0.4                                      # innovation correlation

# covmattransform='z' makes the off-diagonal of a covariance matrix exactly
# atanh(correlation) and the diagonal a standard deviation, so the generating
# values can be written down rather than solved for.
zmat <- function(sd1, sd2, r) matrix(c(sd1, 0, atanh(r), sd2), 2, 2, byrow = TRUE)

# Stationary covariance, solving A S + S A' + Q = 0. Starting T0VAR here makes
# every occasion a draw from the same distribution, so one pair of standard
# deviations describes the whole series.
Q <- diag(diffsd, 2) %*% matrix(c(1, diffcor, diffcor, 1), 2) %*% diag(diffsd, 2)
S <- matrix(solve(kronecker(diag(2), DRIFT) + kronecker(DRIFT, diag(2)),
  -as.vector(Q)), 2, 2)
latentsd <- sqrt(diag(S))

gen <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
  manifestNames = c('eta1', 'eta2'), latentNames = c('eta1', 'eta2'),
  LAMBDA = diag(1, 2), DRIFT = DRIFT,
  DIFFUSION = zmat(diffsd, diffsd, diffcor),
  T0VAR = zmat(latentsd[1], latentsd[2], S[1, 2] / prod(latentsd)),
  T0MEANS = matrix(0, 2, 1), CINT = matrix(0, 2, 1),
  MANIFESTMEANS = matrix(0, 2, 1),
  MANIFESTVAR = diag(1e-6, 2),          # no measurement error: manifest = state
  Tpoints = 30)
gen$covmattransform <- 'z'

# backend='julia' because the R generator reads DIFFUSION and T0VAR as square
# roots whatever covmattransform says, while the julia generator applies the
# declared transform, as ctFit() does.
set.seed(2)
d <- data.frame(ctGenerate(gen, n.subjects = 150, Tpoints = 30, dtmean = 0.7,
  backend = 'julia'))

# The cumulative logit: the response is in category k or below when the state
# plus a logistic disturbance falls below threshold tau_k. Four thresholds give
# five categories.
thresholds <- c(-2.8, -1.0, 1.0, 2.8)
ordinalise <- function(eta)
  1 + rowSums(outer(eta + rlogis(length(eta)), thresholds, '>'))

set.seed(502)
d$y1 <- ordinalise(d$eta1)
d$y2 <- ordinalise(d$eta2)
```

One subject. The line is the latent process, the dotted lines are the
thresholds, and the points are the recorded answers, placed at the middle of
the band each category stands for:

```{r overlay, fig.height = 4}
mid <- c(-3.8, -1.9, 0, 1.9, 3.8)             # band midpoints, for display
s <- d[d$id == 1, ]

ggplot(s, aes(time)) +
  geom_hline(yintercept = thresholds, linetype = 3, colour = 'grey55') +
  geom_line(aes(y = eta1, colour = 'latent eta1')) +
  geom_point(aes(y = mid[y1], colour = 'recorded category y1'), size = 1.9) +
  scale_y_continuous(sec.axis = sec_axis(~., breaks = mid, labels = 1:5,
    name = 'category')) +
  scale_colour_manual(values = c('latent eta1' = 'grey35',
    'recorded category y1' = '#c0392b')) +
  theme_bw() + theme(legend.position = 'bottom') +
  labs(x = 'time', y = 'latent scale', colour = NULL)
```

The points follow the line, but only to the resolution of the bands, and they
cross a threshold slightly early or late depending on the disturbance. Both
items should use the full scale, with no category so rare that it carries
nothing:

```{r categoryuse, fig.height = 2.4}
freq <- rbind(data.frame(item = 'y1', category = factor(d$y1)),
  data.frame(item = 'y2', category = factor(d$y2)))

ggplot(freq, aes(category)) + geom_bar() + facet_wrap(~item) +
  theme_bw() + labs(x = NULL, y = NULL)
```

# What should and should not agree

The two fits place the latent processes on different scales. The Gaussian fit
measures `eta1` in units of the answer codes 1 to 5; the ordinal fit measures
it in logits, because the thresholds fix the scale. Nothing in the data
determines which is right: the latent scale is a convention in both cases.

A parameter can therefore only be compared across the two fits if it is
invariant to rescaling the latent processes.

**Expected to agree**

* *Auto effects* (`drift_eta1`, `drift_eta2`). Their unit is inverse time, and
  a rate of reversion does not change when the process is measured in other
  units.
* *The diffusion correlation*, being a correlation.
* *The standardised cross effect*, the effect expressed in stationary standard
  deviations, $\tilde{a}_{21} = a_{21}\,\text{sd}(\eta_1)/\text{sd}(\eta_2)$.

**Not expected to agree**

* *The raw cross effect* `drift_eta2_eta1`, whose unit is `eta2` per unit of
  `eta1` per unit time. It moves when either scale moves. Note that this is
  already true between the two processes within a single fit: `eta2` is the
  more variable of the two because it accumulates `eta1`, so the same effect
  takes different numbers on differently scaled processes.
* *`DIFFUSION` and `T0VAR` on their own*, and the residual variances, which are
  in latent units or in answer-code units respectively.
* *`MANIFESTMEANS`*, which is a location on the latent scale in the Gaussian
  fit and the first threshold's counterpart in the ordinal one.

Only the ordinal fit's scale has a meaning outside the questionnaire, so where
the raw quantities differ it is the ordinal fit that should be reported.

# Fitting both ways

The two observed-data models differ only in the measurement part: the Gaussian
model estimates a residual variance per item, the ordinal model estimates
thresholds instead.

```{r models}
buildmodel <- function(type, ncat = 0, names = c('y1', 'y2'),
  residual = type != 2) {
  mvar <- diag(0, 2)
  if (residual) diag(mvar) <- paste0('mvar_', names)
  m <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
    manifestNames = names, latentNames = c('eta1', 'eta2'),
    manifesttype = rep(type, 2), ncategories = rep(ncat, 2),
    LAMBDA = diag(1, 2), CINT = matrix(0, 2, 1), T0MEANS = matrix(0, 2, 1),
    MANIFESTVAR = mvar)
  m$pars$indvarying <- FALSE
  m
}

ceilingdata <- setNames(d[, c('id', 'time', 'eta1', 'eta2')],
  c('id', 'time', 'c1', 'c2'))

fits <- list(
  # residual = FALSE because the ceiling data are the latent states themselves:
  # there is no measurement error to estimate, and leaving the parameter free
  # puts it on the boundary at zero, where the likelihood is flat and the fit
  # reports a direction it cannot identify.
  uncategorised = ctFit(ceilingdata,
    buildmodel(0, names = c('c1', 'c2'), residual = FALSE),
    backend = 'julia', cores = 1),
  gaussian = ctFit(d[, c('id', 'time', 'y1', 'y2')], buildmodel(0),
    backend = 'julia', cores = 1),
  ordinal = ctFit(d[, c('id', 'time', 'y1', 'y2')], buildmodel(2, ncat = 5),
    backend = 'julia', cores = 1))
```

`ctDiscretePars()` converts the estimated `DRIFT` into the regressions implied
at each time interval, which is the readable form of the dynamics.
Autoregressions are on the diagonal, cross effects off it:

```{r dynamics, fig.height = 4.6}
ctDiscretePars(fits$ordinal, times = seq(0, 6, .1), standardise = TRUE,
  plot = TRUE, indices = 'all', cores = 1)
```

# Comparison

The standardised cross effect over time, from the three fits and from the
generating parameters:

```{r crosscurve}
times <- seq(0, 6, .1)

crosscurve <- function(fit, label) {
  a <- ctDiscretePars(fit, times = times, standardise = TRUE, cores = 1)
  data.frame(time = times, model = label,
    value = apply(a, c(3, 4, 5), median, na.rm = TRUE)[, 2, 1])
}

curves <- rbind(
  crosscurve(fits$uncategorised, 'uncategorised (ceiling)'),
  crosscurve(fits$gaussian, 'gaussian on the codes'),
  crosscurve(fits$ordinal, 'ordinal'),
  data.frame(time = times, model = 'truth', value = sapply(times,
    function(u) expm::expm(DRIFT * u)[2, 1] * latentsd[1] / latentsd[2])))

ggplot(curves, aes(time, value, colour = model, linetype = model)) +
  geom_line(linewidth = .9) + theme_bw() +
  theme(legend.position = 'bottom') +
  labs(x = 'time interval', y = 'standardised cross effect, eta1 -> eta2',
    colour = NULL, linetype = NULL)
```

The auto effect and the diffusion correlation come from
`summary(fit)$parmatrices`, which contains `DIFFUSIONcov`, the innovation
covariance, and `asymDIFFUSIONcov`, the stationary covariance implied by the
model:

```{r table}
comparable <- function(fit, label) {
  s <- summary(fit)
  pm <- s$parmatrices
  cell <- function(mat, i, j) pm$Mean[pm$matrix == mat & pm$row == i & pm$col == j]
  data.frame(model = label,
    drift_eta1 = s$popmeans['drift_eta1', 'mean'],
    latentsd1 = sqrt(cell('asymDIFFUSIONcov', 1, 1)),
    diffcor = cell('DIFFUSIONcov', 1, 2) /
      sqrt(cell('DIFFUSIONcov', 1, 1) * cell('DIFFUSIONcov', 2, 2)))
}

out <- rbind(
  comparable(fits$uncategorised, 'uncategorised (ceiling)'),
  comparable(fits$gaussian, 'gaussian on the codes'),
  comparable(fits$ordinal, 'ordinal'),
  data.frame(model = 'truth', drift_eta1 = DRIFT[1, 1],
    latentsd1 = latentsd[1], diffcor = diffcor))
cbind(out[1], round(out[-1], 3))
```

`latentsd1` is in answer-code units for the Gaussian fit and in logits for the
ordinal fit, so it is a scale check rather than a comparison.

Categorisation costs little here. That is expected under the assumptions used:
categorisation is a monotone transformation plus noise that is independent
across items and occasions given the latent state, and `MANIFESTVAR` in the
Gaussian model absorbs that noise. The Gaussian model still gets the latent
scale wrong, which is why the raw cross effect cannot be read across the two
fits, but the scale-invariant quantities are recovered.

