---
title: "Treating a rating scale as continuous"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
    self-contained-math: true
execute:
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Treating a rating scale as continuous}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

```{r libs}
library(ctsem)
```

Most longitudinal data in psychology arrive as ratings on five or seven points,
and most models fitted to them treat those ratings as continuous. ctsem will do
either: `manifesttype = 0` puts a Gaussian residual on the observed code and
`manifesttype = 2` puts a cumulative logit on it. This vignette generates data
where the truth is known, fits it both ways, and compares what comes back.

The design is the one where the answer matters --- two processes, one item
each. The parameters at stake are the two such a study reports: the **cross
effect**, how much one process drives the other over time, and the **diffusion
correlation**, how much their innovations move together at an instant. Those
are different claims about the same pair of series, and telling them apart is
most of what a continuous time model is for.

Everything here uses `backend = 'julia'`, because ordinal indicators exist only
on that backend. See the measurement models vignette for the other indicator
types.

# A process we know the truth about

Two Ornstein-Uhlenbeck processes. `eta1` drives `eta2` and not the reverse, so
there is one cross effect to recover rather than two, and their innovations
correlate at 0.4, so there is also a contemporaneous association that is *not*
the cross effect.

```{r truth}
DRIFT <- matrix(c(-0.5, 0,
                  0.3, -0.5), 2, 2, byrow = TRUE)   # eta1 -> eta2 is 0.3
diffsd <- 2.2
diffcor <- 0.4
```

`covmattransform = 'z'` makes a covariance matrix's off-diagonal exactly
`atanh(correlation)`; its diagonal is a standard deviation either way. So the
generating correlation can be written down rather than solved for:

```{r zmat}
zmat <- function(sd1, sd2, r) matrix(c(sd1, 0, atanh(r), sd2), 2, 2,
  byrow = TRUE)
```

Starting from the stationary covariance makes every occasion a draw from the
same distribution, so one pair of standard deviations describes the whole
series. For a linear system that covariance solves
$A\Sigma + \Sigma A^\top + Q = 0$, which is one `solve()` in vectorised form:

```{r stationary}
Q <- diag(diffsd, 2) %*% matrix(c(1, diffcor, diffcor, 1), 2) %*%
  diag(diffsd, 2)
S <- matrix(solve(kronecker(diag(2), DRIFT) + kronecker(DRIFT, diag(2)),
  -as.vector(Q)), 2, 2)
latentsd <- sqrt(diag(S))
latentsd
```

`eta2` ends up more variable because it accumulates `eta1`, and that matters
below: a cross effect's units are *target per source*, so one effect reads as
different numbers on differently scaled processes.

```{r generate}
makelatent <- function(seed, nsubjects = 150, Tpoints = 30, dtmean = 0.7) {
  gen <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
    manifestNames = c('eta1obs', 'eta2obs'), latentNames = c('eta1', 'eta2'),
    LAMBDA = diag(1, 2), DRIFT = DRIFT,
    DIFFUSION = zmat(diffsd, diffsd, diffcor),
    T0VAR = zmat(latentsd[1], latentsd[2], S[1, 2] / prod(latentsd)),
    T0MEANS = matrix(0, 2, 1), CINT = matrix(0, 2, 1),
    MANIFESTMEANS = matrix(0, 2, 1), MANIFESTVAR = diag(1e-6, 2),
    Tpoints = Tpoints)
  gen$covmattransform <- 'z'
  set.seed(seed)
  data.frame(ctGenerate(gen, n.subjects = nsubjects, Tpoints = Tpoints,
    dtmean = dtmean, backend = 'julia'))
}
```

::: {.callout-note}
## `backend = 'julia'` rather than `'r'`

The R generator reads `DIFFUSION` and `T0VAR` as square roots, forming the
covariance as $GG^\top$ whatever `covmattransform` says. The julia generator
applies the transform the model declares, which is what `ctFit()` reads too.
On a univariate model the two agree, a 1x1 matrix having no off-diagonal to
disagree about. Here they do not, so generate from the one that agrees with
the fit.
:::

The measurement step is written out rather than left to `ctGenerate`, so that
the model producing the categories is visibly the one the ordinal fit assumes.
A respondent answers in category $k$ or below with probability
$\text{logit}^{-1}(\tau_k - \eta)$, and four thresholds cut a five point scale:

```{r ordinalise}
invlogit <- function(x) 1 / (1 + exp(-x))

ordinalise <- function(eta, tau) {
  cumulative <- vapply(tau, function(k) invlogit(k - eta), numeric(length(eta)))
  p <- cbind(cumulative, 1)                                # P(y <= k), then 1
  p <- cbind(p[, 1, drop = FALSE], t(apply(p, 1, diff)))   # to P(y = k)
  vapply(seq_len(nrow(p)),
    function(i) sample.int(ncol(p), 1, prob = pmax(p[i, ], 0)), integer(1))
}

thresholds <- c(-2.8, -1.0, 1.0, 2.8)

makedata <- function(seed) {
  d <- makelatent(seed)
  set.seed(seed + 500)
  d$y1 <- ordinalise(d$eta1obs, thresholds)
  d$y2 <- ordinalise(d$eta2obs, thresholds)
  d
}

d <- makedata(2)
rbind(y1 = round(prop.table(table(d$y1)), 3),
  y2 = round(prop.table(table(d$y2)), 3))
```

One item per process, both using the whole scale, with no category rare enough
to carry nothing.

# What can be compared, and what cannot

The two fits put their latents on different rulers: the Gaussian fit measures
`eta1` in units of y1's answer codes, the ordinal fit in logits. A parameter is
comparable across the two exactly when it survives rescaling the latents.

**Auto effects survive** --- `drift_eta1` has units of inverse time.

**The raw cross effect does not.** `drift_eta2_eta1` is in units of `eta2` per
unit of `eta1` per unit time, so it moves if either ruler moves. What survives
is the standardised version, the effect in stationary standard deviations:

$$
\tilde{a}_{21} \;=\; a_{21}\,\frac{\text{sd}(\eta_1)}{\text{sd}(\eta_2)}
$$

**The diffusion correlation survives**, being a correlation.

```{r targets}
truth <- data.frame(model = 'truth',
  drift_eta1 = DRIFT[1, 1], drift_eta2 = DRIFT[2, 2],
  cross_raw = DRIFT[2, 1],
  cross_std = DRIFT[2, 1] * latentsd[1] / latentsd[2],
  diffcor = diffcor, row.names = NULL)
truth
```

# The fits

The models differ in three lines: the Gaussian one puts a free residual
variance on each observed code, the ordinal one has none to put and estimates
thresholds instead.

```{r models}
buildmodel <- function(type, ncat = 0, names = c('y1', 'y2')) {
  mvar <- diag(0, 2)
  if (type != 2) {
    mvar[1, 1] <- paste0('mvar_', names[1])
    mvar[2, 2] <- paste0('mvar_', names[2])
  }
  m <- ctModel(type = 'ct', n.latent = 2, n.manifest = 2,
    manifestNames = names, latentNames = c('eta1', 'eta2'),
    manifesttype = rep(type, 2), ncategories = rep(ncat, 2),
    LAMBDA = diag(1, 2), CINT = matrix(0, 2, 1), T0MEANS = matrix(0, 2, 1),
    MANIFESTVAR = mvar)
  m$pars$indvarying <- FALSE
  m
}
```

Alongside the two fits of interest there is a third, and it is what makes the
other two readable: the same model fitted to the **uncategorised** latent
states. It is not available with real data, and it is the ceiling --- whatever
it misses belongs to the design and the sample rather than to categorising.

```{r fits}
fitall <- function(d) {
  ceiling <- setNames(d[, c('id', 'time', 'eta1obs', 'eta2obs')],
    c('id', 'time', 'c1', 'c2'))
  list(
    uncategorised = ctFit(ceiling, buildmodel(0, names = c('c1', 'c2')),
      backend = 'julia', cores = 1),
    gaussian = ctFit(d[, c('id', 'time', 'y1', 'y2')], buildmodel(0),
      backend = 'julia', cores = 1),
    ordinal = ctFit(d[, c('id', 'time', 'y1', 'y2')],
      buildmodel(2, ncat = 5), backend = 'julia', cores = 1))
}
fits <- fitall(d)
```

`summary()` carries `DIFFUSIONcov` and `asymDIFFUSIONcov` in `parmatrices` ---
the innovation covariance and the stationary covariance the model implies. The
first gives the diffusion correlation, the second the standard deviations that
standardise the cross effect.

```{r extract}
readfit <- function(fit, label) {
  e <- summary(fit)$popmeans
  pm <- summary(fit)$parmatrices
  cell <- function(mat, i, j)
    pm$Mean[pm$matrix == mat & pm$row == i & pm$col == j]
  sds <- sqrt(c(cell('asymDIFFUSIONcov', 1, 1), cell('asymDIFFUSIONcov', 2, 2)))
  data.frame(model = label,
    drift_eta1 = e['drift_eta1', 'mean'],
    drift_eta2 = e['drift_eta2', 'mean'],
    cross_raw = e['drift_eta2_eta1', 'mean'],
    cross_std = e['drift_eta2_eta1', 'mean'] * sds[1] / sds[2],
    diffcor = cell('DIFFUSIONcov', 1, 2) /
      sqrt(cell('DIFFUSIONcov', 1, 1) * cell('DIFFUSIONcov', 2, 2)),
    row.names = NULL)
}

readall <- function(fits) {
  out <- rbind(readfit(fits$uncategorised, 'uncategorised (ceiling)'),
    readfit(fits$gaussian, 'gaussian on the codes'),
    readfit(fits$ordinal, 'ordinal'), truth)
  cbind(out[1], round(out[-1], 3))
}
readall(fits)
```

Standard errors on the two the summary reports directly:

```{r ses}
data.frame(model = names(fits),
  drift_eta1_se = sapply(fits, function(f)
    summary(f)$popmeans['drift_eta1', 'sd']),
  cross_raw_se = sapply(fits, function(f)
    summary(f)$popmeans['drift_eta2_eta1', 'sd']),
  row.names = NULL)
```

**Categorising cost almost nothing here.** The standardised cross effect comes
back at 0.225 from the Gaussian fit and 0.219 from the ordinal one, against
0.196 from the ceiling fit that never saw a category; the diffusion correlation
at 0.388 and 0.396 against the ceiling's 0.416. The gaps between the three
fits --- 0.02 to 0.03 on both quantities --- are no larger than the gap between
the ceiling fit and the generating 0.252 and 0.400, which is to say that most
of the distance from truth in this table is sampling variability in the
processes themselves rather than anything the measurement model did. The auto
effects come back within a standard error either way.

That is not the answer the question usually expects, and the reason is worth
stating: categorising is a monotone transformation plus noise that is
independent across items and occasions given the state, and `MANIFESTVAR` is
already there to absorb the noise. The Gaussian model gets the arbitrary latent
scale wrong, which is why `cross_raw` should not be read across the two rows ---
but the comparable quantities survive.

# A second dataset

One dataset is one draw. The same design and a different seed:

```{r second}
d2 <- makedata(1)
fits2 <- fitall(d2)
readall(fits2)
```

The standardised cross effect holds up again --- 0.269 and 0.270 against the
ceiling's 0.274 --- and this time it is the ordinal fit that misbehaves
elsewhere, returning `drift_eta1` at -1.099 and a diffusion correlation of
0.256 where the Gaussian fit gives -0.521 and 0.453.

::: {.callout-warning}
## One ordinal item per process: check the implied scale

That second-dataset `drift_eta1` is not sampling noise, and it is worth knowing
about before it happens to you. With a single ordinal item on a process, the
item's reliability and the speed of the process trade against each other: a
latent twice as variable makes the item twice as discriminating, and a process
reverting twice as fast undoes the extra agreement between successive answers,
so the observed sequence of categories is nearly the same either way. Two items
on one latent break the trade at once, through their agreement at a single
occasion; one item leaves only the shape of the autocorrelation to do it.

The signature is an auto effect and a `DIFFUSION` far from anything plausible,
moving together, while the standardised quantities stay put --- exactly the
second table. The standard errors do not flag it, because the ridge is curved
and curvature at the optimum says nothing about how far along it that optimum
sits. Check `sqrt(diag(asymDIFFUSIONcov))` against the range the item can
actually resolve before reading the dynamics, and prefer an item that
discriminates across that range: on a version of this design with the process
standard deviation at 1.3 against thresholds at $\pm 1.6$, every seed tried
went this way.
:::

# Summary

* On a five point scale used across its range, treating it as continuous cost
  little here: the standardised cross effect and the diffusion correlation came
  back from the Gaussian fit about as close to the ceiling fit as the ordinal
  fit did, on both datasets.
* Compare **standardised** cross effects across the two. The raw ones are in
  units of two different rulers, and only the ordinal fit's ruler means
  anything outside the questionnaire.
* `summary(fit)$parmatrices` carries `DIFFUSIONcov` and `asymDIFFUSIONcov`,
  which is where the diffusion correlation and the standardising standard
  deviations come from.
* Fit the same model to the uncategorised states whenever the data are
  simulated. Anything the ceiling fit misses belongs to the design, not to
  categorising.
* With one ordinal item per process, check the fit's implied latent standard
  deviation before reading `DIFFUSION` or an auto effect from it.
* Generate with `backend = 'julia'` whenever a covariance has an off-diagonal:
  the R generator reads `DIFFUSION` and `T0VAR` as square roots regardless of
  `covmattransform`, and the julia one reads them as `ctFit()` does.
